### 2013-11-30 paper/paper.tex
Added new resources for compiling final report. View README instructions inside the paper folder for how to build the paper.

%\documentclass[3p]{elsarticle}
\documentclass[5p]{elsarticle}
\usepackage{natbib,graphicx,amsmath, mhchem,url,colortbl,fancyhdr,subfigure}
\usepackage{multicol}

\journal{EECS 545}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}

\begin{document}
\begin{frontmatter}

%\input{header.tex}
\title{Methods for Polyphonic Music Transcription} % Remove Piano part?
\author[add]{Jeremy Nash}
\ead{nashj@umich.edu}
\author[add]{Mark Liu}
\ead{markmliu@umich.edu}
\author[add]{Paul Schroeder}
\ead{pschro@umich.edu}
\address[add]{Electrical Engineering Department, University of Michigan, Ann Arbor, MI 48109}

%\input{abstract.tex}
\begin{abstract}
This paper presents a review of modern methods for polyphonic music transcription. NNN
\end{abstract}
\begin{keyword}
Transcription \sep sparse coding \sep non-negative matrix factorization \sep Bayesian non-parametrics \sep music information retrieval
\end{keyword}

\end{frontmatter}

\setcounter{topnumber}{1}

\section{Motivation}
% Explain like you're at a whiteboard

% Here's why people want to do this
% Study of music - understanding music
% recommender systems
% Music information database
% For learning musicians

% Here's why it's difficult
% Musical timbres
% Time varying - attack, decay, sustain, release
% Number of sources?
% Human perception of pitch
% Musical instrument tuning
% Completely new instruments - software created

% Here's a teaser on how I'm going to solve this problem

% Here's what I'm about to show you

\section{Problem Statement}

\section{Related Work}
% Contain a table with metrics data, brief description, year published

\section{Methodologies}

The two main methodologies in modern polyphonic music transcription reflect the supervised/unsupervised learning dichotomy in machine learning. The first approach learns separate discriminative models for the presence of a note over spectral features.

% Constant Q transform

% Semitone filter bank

\subsection{Spectrogram}
% Hint at the matrix factorization approach
\subsection{Constant Q transform}
\subsection{Semitone filter bank}
\subsection{Discriminative models}
% SVM, logistic regression
\subsection{SVM}
\subsection{Feed forward neural network}

\subsection{Non-negative matrix factorization}
% Non-negative Matrix factorization
\subsection{Bayesian nonparametric models}
\subsection{Smoothing}
\subsubsection{HMM smoothing}
\subsubsection{Probabilistic spectral smoothness}
\subsection{Recurrent neural network}

\section{Evaluation}

\section{Conclusion}
% Which method is fastest

% Which method has the best performance

% Which one is the simplest method

% Future directions

\section{Individual Effort}
\begin{description}
\item[$\bullet$ Jeremy] wrote and evaluted the SVM method in \cite{poliner2006discriminative}, the Bayesian nonparametric method in \cite{blei2010bayesian}, and the sparse non-negative matrix factorization method in \cite{abdallah2004polyphonic}. Jeremy also wrote the paper.
\item[$\bullet$ Mark] implemented the hidden Markov model in \cite{poliner2006discriminative} and the LSTM network in \cite{bock2012polyphonic}.
\item[$\bullet$ Paul] implemented the LSTM network in \cite{bock2012polyphonic}.
\end{description}

%    Problem statement, both conceptual and mathematical.
%    Motivation: why is the problem important.
%    Related work.
%    Methodologies explored and developed, and their advantages and disadvantages
%    Evaluation: experiments on sythetic and real data, performance measures used, results. I hope to see quantitative, objective evaluations, and comparisons with competitors
%    Conclusions: what did you learn, what were the project's success and failures?
%    Description of individual effort: At the end of the report, in a paragraph (included in the page limit), please include a brief description of each project member's contribution to the project.

% Support vector machines + HMM

% Honglak Lee's approach: use knn or RBM for input

% Non-negative matrix factorization

% Supervised NMF

% Combination NMF + svm

% Recurrent neural network (LSTM)

% Bayesian non-parametric method

% Bayesian non-parametric II (infinite hidden markov model)

% Our approach: Bayesian non parametrics with stochastic ...

%\input{1.tex}
%\input{fig1.tex}
%\input{2.tex}
%\input{3.tex}
%\input{fig2.tex}
%\input{4.tex}
%\input{5.tex}

% Your project should involve experiments that compare a few different algorithms. These should be well known or state-of-the-art methods, plus possibly any new methods you develop. The experiments should use real (as opposed to simulated) data sets. There is a large repository of machine learning data sets called the ``UCI Machine Learning Repository.'' There are other more domain-specific sources as well that you might find by searching.

%Your project should focus on a topic that has previously been studied in the machine learning literature. I include this requirement because in the past, some groups have defined entirely new problems for themselves, and been unable to make any progress.
%Your project report should contain an extensive literature review that summarizes fundamental ideas and state-of-the-art methods, with critiques of their strengths and weaknesses. My favorite reports to read are the ones that teach me something interesting.
%You are also encouraged to develop your own methods to solve your chosen problem. 
%Your project should involve experiments that compare a few different algorithms. These should be well known or state-of-the-art methods, plus possibly any new methods you develop. The experiments should use real (as opposed to simulated) data sets. There is a large repository of machine learning data sets called the ``UCI Machine Learning Repository.'' There are other more domain-specific sources as well that you might find by searching.

%Every report should address the following (but don't feel obligated to make these the section headings)
%    Problem statement, both conceptual and mathematical.
%    Motivation: why is the problem important.
%    Related work.
%    Methodologies explored and developed, and their advantages and disadvantages
%    Evaluation: experiments on sythetic and real data, performance measures used, results. I hope to see quantitative, objective evaluations, and comparisons with competitors
%    Conclusions: what did you learn, what were the project's success and failures?
%    Description of individual effort: At the end of the report, in a paragraph (included in the page limit), please include a brief description of each project member's contribution to the project.

\bibliographystyle{model1-num-names}
\bibliography{transcription}
\end{document}

### 2013-11-30 paper/maketex.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+pdflatex paper.tex
+bibtex paper
+pdflatex paper.tex
+pdflatex paper.tex

### 2013-12-01 paper/paper.tex
Added an introduction, some notes on NMF and frequency scales, evaluation, and individual efforts to the paper

@@
 \title{Methods for Polyphonic Music Transcription} % Remove Piano part?
-\author[add]{Jeremy Nash}
+% 'Everything you wanted to know about polyphonic music transcription' 
+% Note: according to Scott's final report list, it looks like the best paper according to him is a really thorough survey of an already established field, and he repeatedly indicates that he doesn't necessarily expect any new methods
+\author[eadd]{Jeremy Nash}
 \ead{nashj@umich.edu}
-\author[add]{Mark Liu}
+\author[eadd]{Paul Schroeder}
+\ead{pschro@umich.edu}
+\address[eadd]{Electrical Engineering Department, University of Michigan, Ann Arbor, MI 48109}
+\author[csadd]{Mark Liu}
 \ead{markmliu@umich.edu}
-\author[add]{Paul Schroeder}
-\ead{pschro@umich.edu}
-\address[add]{Electrical Engineering Department, University of Michigan, Ann Arbor, MI 48109}
+\address[csadd]{Computer Science Department, University of Michigan, Ann Arbor, MI 48109}
 
 %\input{abstract.tex}
+
 \begin{abstract}
-This paper presents a review of modern methods for polyphonic music transcription. NNN
+% 4 sentences - write this last
+
+% State the problem
+We present a review of modern methods for polyphonic music transcription.
+
+% Say why it's an interesting problem
+
+% Say what your solution achieves
+
+% Say what follows from your solution
+
 \end{abstract}
@@
 
+\section{Introduction}
+
+\subsection{Terminology}
+Semitones are the smallest music interval in Western music. On a piano, semitones are the adjacent keys.
+
+Timbre represents the spectral envelope, time envelope, noise, tonality, and onset of a musical note. Timbre captures the unique properties of a musical instrument.
+
+Pitches are represented by the notes used in musical notation.
+Pitches are a psychological perception and not an objective physical property.
+While we can intuitively assign a linear order to pitches (by referring to pitches as lower or higher than other pitches), pitches are rarely a single frequency and are instead a function of their timbre.
+
 \section{Motivation}
+% No more than one page
+
+% Describe the problem 
+%    Use an example to introduce the problem
+% State your contributions
+%    Bulleted list of contributions
+
 % Explain like you're at a whiteboard
@@
 
+% According to Simon Peyton Jones:
+% Here is a problem
+% It's an interesting problem
+% It's an unsolved problem
+% Here is my idea
+% My idea works 
+% Here's how my idea compares to other approaches
+
 \section{Problem Statement}
+% No more than one page
 
@@
 \section{Methodologies}
+There are currently two main approaches to polyphonic music transcription.
+The first approach learns a supervised learning model that uses frequency magnitudes to predict the presence of a note.
+The second approach uses unsupervised learning techniques to extract
 
-The two main methodologies in modern polyphonic music transcription reflect the supervised/unsupervised learning dichotomy in machine learning.
-The first approach learns separate discriminative models for the presence of a note over spectral features.
+\subsection{Audio data}
 
-% Constant Q transform
+% Sampling rate for piano
 
-% Semitone filter bank
-
+% STFT
 \subsection{Spectrogram}
@@
 \subsection{Constant Q transform}
+
 \subsection{Semitone filter bank}
+Unlike the linear frequency scale used in an FFT, semitones in Western music are spaced logarithmically.
+This logarithmic spacing derives from our roughly logarithmic perception of pitches, first captured by \citet{stevens1937scale} through the mel scale.
+
+Transforming the spectra to a logarithmic frequency scale achieves two main benefits.
+It reduces the dimensionality of the training data, since most of the spectral energy occupies a logarithmic grid of frequencies.
+In practice, \citet{bock2012polyphonic} were able to reduce the dimensionality of their training data from 5120 FFT magnitudes down to 183.
+
+The logarithmic frequency scale also decreases sensitivty to detuning.
+In a linear frequency scale, detuning, especially in higher semitones, will cause spectral energy to leak into neighboring frequency bins, creating more variability in the training data.
+On a logarithmic frequency scale, minor detuning would result in less spectral energy leakage and less unnecessary variability in the training data.
+
+\subsection{Note onset detection}
+This method proposes restricting the problem to determining note onsets.
+This is a simpler problem than detecting note durations because of the possibility for time-varying timbres in instruments.
+Most researchers in this field evaluate their music transcription methods on their note onset detection performance.
 \subsection{Discriminative models}
@@
 \subsection{SVM}
+% Frequency partition
+
 \subsection{Feed forward neural network}
@@
 \subsection{Non-negative matrix factorization}
+
+\citet{lee1999learning}
+\citet{seung2001algorithms}
+
+\subsection{Sparse non-negative matrix factorization}
+
+One of the great side-effects of NMF is that it often produces a sparse, parts-based representation.
+This is desirable in music transcription because the piano roll matrix, $H$, is a sparse matrix of note activations.
+However, the NMF method does not contain an explicit sparsity objective.
+\citet{hoyer2004non} extended NMF to contain a tunable sparsity objective.
+
+The sparse NMF method in \citet{hoyer2004non} was adopted by \citet{abdallah2004polyphonic} in their transcription system.
+
 % Non-negative Matrix factorization
@@
 \section{Evaluation}
+
+There is no standard metric to evaluate the performance of a transcription method. However, many researchers use the following methods:
+
+\subsection{MIDI databases}
+
+\subsection{Evaluation measures}
+
+\subsubsection{}
+
+\subsubsection{Frame-level transcription error score}
+
+\subsection{Method comparison}
+
+\subsubsection{Reported measures}
+\begin{center}
+\begin{tabular}{|l|l|l|}
+\hline
+Algorithm & Accuracy & Score \\ \hline
+SVM & 21\% & 21 \\ \hline
+SVM with HMM & 22\% & 22 \\ \hline
+NMF & 23\% & 23 \\ \hline
+Sparse NMF & 24\% & 24 \\ \hline
+\end{tabular}
+\end{center}
+
+\subsubsection{Our evaluation}
+
+The following table captures our evaluation of these methods.
+We expect these accuracy measures to be lower than they were reported in the original papers because we are using a smaller training dataset and not attempting to optimize the methods.
+
+\begin{center}
+\begin{tabular}{|l|l|l|}
+\hline
+Algorithm & Accuracy & Score \\ \hline
+SVM & 18\% & 18 \\ \hline
+SVM with HMM & 19\% & 19 \\ \hline
+NMF & 20\% & 20 \\ \hline
+Sparse NMF & 21\% & 21 \\ \hline
+\end{tabular}
+\end{center}
 
@@
 \begin{description}
-\item[$\bullet$ Jeremy] wrote and evaluted the SVM method in \cite{poliner2006discriminative}, the Bayesian nonparametric method in \cite{blei2010bayesian}, and the sparse non-negative matrix factorization method in \cite{abdallah2004polyphonic}.
+\item[$\bullet$ Jeremy] wrote and evaluted the SVM method in \citet{poliner2006discriminative}, the Bayesian nonparametric method in \citet{blei2010bayesian}, and the sparse non-negative matrix factorization method in \citet{abdallah2004polyphonic}.
 Jeremy also wrote the paper.
-\item[$\bullet$ Mark] implemented the hidden Markov model in \cite{poliner2006discriminative} and the LSTM network in \cite{bock2012polyphonic}.
-\item[$\bullet$ Paul] implemented the LSTM network in \cite{bock2012polyphonic}.
+\item[$\bullet$ Mark] implemented the hidden Markov model in \citet{poliner2006discriminative} and the LSTM network in \citet{bock2012polyphonic}.
+\item[$\bullet$ Paul] implemented the LSTM network in \citet{bock2012polyphonic}.
 \end{description}

### 2013-12-01 paper/paper.tex
Added introduction, terminology, motivation, and some methodologies like NMF

@@
 % State the problem
-We present a review of modern methods for polyphonic music transcription.
+%We present a review of modern methods for polyphonic music transcription.
+
+Music transcription is the process of converting an audio recording into a musical score. 
+Musicians have been transcribing music manually for centuries to learn new songs and gain musical insight.
+Transcription, however, is an arduous manual process for musicians, and it's a difficult problem for computers because the pitches in a musical score are the human perceptions of a time-varying and frequency-varying physical phenomenon.
+There has been a flurry of research on how to solve this problem. 
+We investigate the best solutions to the automated transcription problem and evaluate and compare their performance.
 
@@
 
+Polyphony means multiple simultaneous notes. Monophony means only a single note at a time. These terms carry slightly different meanings outside of the automated transcription field.
+
 \section{Motivation}
 % No more than one page
+
+% Introduction
+
+% Explain motivation
+
+% This is just copied from 'A Connectionist Approach to Automatic Transcription of Polyphonic Piano Music' for now:
+Applications of a music transcription are versatile.
+Transcription produces a compact and standardized parametric representation of music.
+Such representation is needed for content-based retrieval of music in most current musical databases.
+It is useful in music analysis systems for tasks such as melody extraction, music segmentation and rhythm tracking.
+Transcription aids musicologists in analyzing music that has never been written down, such as improvised or ethnical music.
+The conversion of an acoustical waveform into parametric description is also useful in the process of making music, as well as in newer coding standards, such as MPEG-4, which may include such descriptions.
+
+% According to Simon Peyton Jones:
+% Here is a problem
+% It's an interesting problem
+% It's an unsolved problem
+% Here is my idea
+% My idea works 
+% Here's how my idea compares to other approaches
 
@@
 % Here's what I'm about to show you
-
-% According to Simon Peyton Jones:
-% Here is a problem
-% It's an interesting problem
-% It's an unsolved problem
-% Here is my idea
-% My idea works 
-% Here's how my idea compares to other approaches
 
@@
 \subsection{Non-negative matrix factorization}
+
+Non-negative matrix factorization (NMF) is an unsupervised, dimensionality reduction technique that factors a non-negative matrix $V$ with dimensions $FxN$ into two matrices,
+
+\[ V \approx WH \]
+
+\noindent where $W$ is an $FxK$ matrix and $H$ is a $KxN$ matrix. $K$ is typically chosen so that $F*K + K*N << F*N$.
+
+In piano transcription, for example, $V$ might be a spectrogram with dimensions $2048xN$, $W$ might be $2048x88$ and encode the spectral envelope for each piano key in each column of the matrix, and $H$ might be $88xN$ sparse piano roll matrix.
+
+There are two cost functions commonly used for NMF. The first is a euclidean distance cost function, defined as the squared difference between the corresponding elements in each matrix:
+
+\[ \displaystyle ||V-WH||^{2} = \sum_{ij} (V_{ij} - (WH)_{ij})^{2} \]
+
+\noindent Another commonly used measure is the KL divergence:
+
+\[ \displaystyle D(V||WH) = \sum_{ij} (V_{ij} \log \frac{V_{ij}}{(WH)_{ij}} - V_{ij} + (WH)_{ij}) \]
+
+The KL divergence cost function is more often used by transcription researchers because it XXX.
 

### 2013-12-02 paper/paper.tex
Replaced letter x with 'times' symbol when describing matrix sizes in the non-negative matrix factorization section

@@
 
-Non-negative matrix factorization (NMF) is an unsupervised, dimensionality reduction technique that factors a non-negative matrix $V$ with dimensions $FxN$ into two matrices,
+Non-negative matrix factorization (NMF) is an unsupervised, dimensionality reduction technique that factors a non-negative matrix $V$ with dimensions $F \times N$ into two matrices,
 
@@
 
-\noindent where $W$ is an $FxK$ matrix and $H$ is a $KxN$ matrix. $K$ is typically chosen so that $F*K + K*N << F*N$.
+\noindent where $W$ is an $F \times K$ matrix and $H$ is a $K \times N$ matrix. $K$ is typically chosen so that $F*K + K*N << F*N$.
 

### 2013-12-03 bin/poliner_svm.m
Created separate .m files for the SVM method and NMF method

@@ -1,4 +1,4 @@
-function [estimated_pr, pr] = transcribe(wav_file, midi_file)
+function [estimated_pr, pr] = poliner_svm(wav_file, midi_file)
   % Warning: this is written for Octave and not tested with MATLAB
   % This also uses libsvm. The libsvm I have included has been built for Octave
 
@@ -51,6 +51,10 @@ function [estimated_pr, pr] = transcribe(wav_file, midi_file)
   notes = midiInfo(midi,0);
   % T is in increments of 10ms, the same as our STFT'd wav file
   [pr, t, nn] = piano_roll(notes);
+  notes
+  pause
+  
+
   size(pr)
   % pause
   % sum(pr,2)
@@ -163,42 +167,46 @@ function [estimated_pr, pr] = transcribe(wav_file, midi_file)
 
   % Run a hidden markov model over the piano roll for smoothing the raw log posterior probabilities
   % Estimation of priors and transition matrix can be done using MIDI files (ground truth)
-  % Estimation of emission matrix should be done using predict_label from
-  % SVM
+  % Estimation of emission matrix should be done using predict_label from SVM
   
   notes_list = unique(notes(:,3)); %list of notes that are played
   true_labels = zeros(size(estimated_pr));
   smooth_labels = zeros(size(estimated_pr));
   true_labels = bsxfun(@minus,true_labels,1); %initialize true_labels to -1
+
   for i=1:size(nn,2)
     fprintf('Smoothing note %d\n', i);
     cur_note = nn(i);
+
   % If cur_note is not actually played in song, skip it
-    if (isempty(find(notes_list==cur_note, 1)))
+    if (isempty(find(notes_list == cur_note, 1)))
         continue;
     end
+
   % Estimate prior
-    on_times = notes(notes(:,3)==cur_note,:);
+    on_times = notes(notes(:,3) == cur_note,:);
     on_time_total = 0;
-    time_total = t(size(t,2)); %using this as proxy for total length of time
+    time_total = t(end); %using this as proxy for total length of time
     for j = 1:size(on_times,1)
         %sum the differences between on and off time of note to get total
         %on_time
-        on_time_total=on_time_total+(on_times(j,6)-on_times(j,5));
+        on_time_total = on_time_total + (on_times(j,6) - on_times(j,5));
     end
[…]

### 2013-12-03 bin/smaragdis_nmf.m

@@ -1,0 +1,92 @@
+function [estimated_pr, pr, H, W] = smaragdis_nmf(wav_file, midi_file)
+  % Warning: this is written for Octave and not tested with MATLAB
+
+  % ***Note to self***
+  % I'm worried that the wav file and midi file are not aligned properly
+
+  % Setup
+  % 0. Generate 16KHz wave file from midi
+  
+  % Testing
+  % 1. Compute STFT of wave file
+  % 2. Run NMF on the spectrogram, starting with random matrices
+  % 3. Apply a threshold to the piano roll
+
+  more off;
+
+  % Open wav file
+  [y,fs,bps] = wavread(wav_file);
+  % Compute the STFT on 128ms frames with 10ms hops
+  isOctave = exist('OCTAVE_VERSION') ~= 0;
+  if isOctave
+    [S, f, spec_t] = specgram(y, 2048, fs, hanning(.128*fs), 2048 - 160); % Needs to be spectrogram in MATLAB  
+  else
+    [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
+  end
+
+  % Compute the FFT magnitudes
+  magS = abs(S);
+
+  magS(1:40,1:40)
+  pause
+
+  num_steps = size(S,2);  
+
+  % Add the midi library
+  addpath('../lib/matlab-midi/src');
+  % Open the MIDI file
+  midi = readmidi(midi_file);
+  notes = midiInfo(midi,0);
+  % T is in increments of 10ms, the same as our STFT'd wav file
+  [pr, t, nn] = piano_roll(notes);
+
+  % Only display a subset of the ground truth for comparison to computed piano roll later
+  subset = 2000;
+  view_piano_roll(t(1:subset),nn,pr(:,1:subset), 'Ground truth')
+  
+  % Run NMF on the spectrogram
+  
+  % Randomly generate W and H
+  W = rand(1024, 88);
+  %H = rand(88, num_steps);
+
+  % Perform on subset 
+  magS = magS(:,1:subset);
+  H = rand(88, subset);
+  
+  % Iterate on W and H until convergence
+  for i=1:1000 % Use actual stopping criteria here
+      fprintf('Update #%d:\n', i);
+      
+      % Apply update rule to W and H
+      %Wnew = W + (W*H - magS) * H';
+      %Hnew = H + W'*(W*H - magS);
[…]

### 2013-12-03 bin/smaragdis_nmf.m
Fixed the update rules and set notes (k) = 25 for now for smaragdis_nmf. The method actually produces a sparse solution now but is difficult to interpret because the notes are unordered.

@@ -27,8 +27,8 @@ function [estimated_pr, pr, H, W] = smaragdis_nmf(wav_file, midi_file)
   % Compute the FFT magnitudes
   magS = abs(S);
 
-  magS(1:40,1:40)
-  pause
+  %magS(1:40,1:40)
+  %pause
 
   num_steps = size(S,2);  
 
@@ -47,38 +47,37 @@ function [estimated_pr, pr, H, W] = smaragdis_nmf(wav_file, midi_file)
   % Run NMF on the spectrogram
   
   % Randomly generate W and H
-  W = rand(1024, 88);
-  %H = rand(88, num_steps);
+  k = 25;
+  W = rand(1024, k);
+  %H = rand(k, num_steps);
 
   % Perform on subset 
   magS = magS(:,1:subset);
-  H = rand(88, subset);
+  H = rand(k, subset);
   
+  V = magS; 
+
   % Iterate on W and H until convergence
-  for i=1:1000 % Use actual stopping criteria here
+  for i=1:200 % Use actual stopping criteria here
       fprintf('Update #%d:\n', i);
       
       % Apply update rule to W and H
       %Wnew = W + (W*H - magS) * H';
       %Hnew = H + W'*(W*H - magS);
-      Hnew = H + ((W'*magS) ./ (W'*W*H));
-      Wnew = W + ((magS*H') ./ (W*H*H'));
-
+      H = H .* ((W'*V) ./ (W'*W*H + 1e-9));
+      W = W .* ((V*H') ./ (W*H*H' + 1e-9));
+ 
       % Compute the Frobenius norm
-      H_norm = norm(H-Hnew,'fro');
-      W_norm = norm(W-Wnew,'fro');
-      fprintf('H_norm: %g\nW_norm: %g\n\n', H_norm, W_norm);
-
-      H = Hnew;
-      W = Wnew;
+      S_norm = norm(V - W*H, 'fro'); % Shows how close W*H is to the spectrogram
+      fprintf('S_norm: %g\n\n', S_norm);
   end
   
   % Threshold the values in the H piano roll matrix
 
+  eta = .1
   estimated_pr = zeros(size(pr,1), size(pr,2));
-  eta = .75;
   H = H ./ max(max(H)); % Adding this to make the threshold meaningful
-  for i=1:88
+  for i=1:k
       for j=1:subset
       	  if (H(i,j) > eta)
 	     estimated_pr(i,j) = 1;

### 2013-12-03 bin/view_piano_roll.m
Added view_piano_roll function

@@ -1,0 +1,15 @@
+function view_piano_roll(t, nn, pr, title_str)
+  % View piano roll
+  % t is an array of times, usually in increments of .01s
+  % nn are the note numbers 
+  % pr is the piano roll binary matrix
+  
+  figure;
+  imagesc(t,nn,pr);
+  axis xy;
+  title(title_str);
+  xlabel('Time');
+  ylabel('Notes');
+  % pause;
+  drawnow
+end

### 2013-12-04 bin/const_q.m
Added constant-Q functions and demo

@@ -1,0 +1,10 @@
+function cq = const_q(x, minFreq, maxFreq, bins, fs)
+  Q= 1/(2^(1/bins)-1);
+  K = ceil(bins*log2(maxFreq/minFreq));
+  cq = zeros(1,K);
+  for k=1:K
+    fk = minFreq*2^((k-1)/bins);
+    N = round(Q*fs/fk);
+    cq(k) = x(1:N) * (hamming(N) .* exp(-2*pi*i*Q*(0:N-1)'/N))/N;
+  end
+end

### 2013-12-04 bin/const_q_demo.m

@@ -1,0 +1,29 @@
+function Q = const_q_demo
+
+more off;
+
+[y,fs,bps] = wavread('../data/br_im2.wav');
+length(y)
+% Every 10 ms = 160 samples shift
+steps = (length(y) - 19856)/160 + 1;
+spec = 175;
+% spec = 88;
+Q = zeros(spec, steps);
+step_range = 1:3000;
+sparkernel = sparse_kernel(27.5, 4186.01, 24, 16000);
+
+for i=step_range
+   fprintf('Iter: %d\n', i);
+   %q = const_q(y(1+(i-1)*160:end)', 27.5, 4186.01, 12, 16000);
+   q = const_q_fast(y(1+(i-1)*160:end)', sparkernel);
+   Q(:,i) = q;
+end
+
+Q = abs(Q(:, step_range));
+imagesc(Q);
+pause
+
+end
+
+
+

### 2013-12-04 bin/const_q_fast.m

@@ -1,0 +1,6 @@
+function cq = const_q_fast(x, sparKernel)    % x must be a row vector
+  cq = fft(x,size(sparKernel,1)) * sparKernel;   
+end
+
+% Note: sparKernel takes a couple seconds to compute, but only needs to be computed once. This is why it's added in separately
+% See const_q_demo for an example

### 2013-12-04 bin/sparse_kernel.m

@@ -1,0 +1,21 @@
+function sparKernel= sparse_kernel(minFreq, maxFreq, bins, fs, thresh)
+% This is used for the fast constant-Q algorithm 
+
+if nargin<5 thresh = 0.0054; end    % for Hamming window
+Q = 1/(2^(1/bins)-1);                                                    
+K = ceil( bins * log2(maxFreq/minFreq) );                                
+fftLen = 2^nextpow2( ceil(Q*fs/minFreq) );
+tempKernel = zeros(fftLen, 1);
+sparKernel = [];
+for k = K:-1:1
+   len = ceil( Q * fs / (minFreq*2^((k-1)/bins)) );                      
+   tempKernel(1:len) = hamming(len)/len .* exp(2*pi*i*Q*(0:len-1)'/len); 
+   specKernel = fft(tempKernel);                                         
+   specKernel(find(abs(specKernel)<=thresh)) = 0;
+   sparKernel = sparse([specKernel sparKernel]);
+end
+sparKernel = conj(sparKernel) / fftLen;                                  
+
+end
+
+

### 2013-12-04 bin/qgram.m
Added a constant-q spectrogram function called qgram. It takes a few seconds to run, but it is much faster than the old constant q function.

@@ -1,0 +1,38 @@
+function Q = qgram(y, fs, bins_multiplier)
+% y is the signal
+% fs is the sampling frequency
+% bins_multiplier multiples the number of bins per octave. When bins_multiplier=1, bins=12 (which is the number of semitones per octave). It seems like researchers typically choose bins_multiplier=2 for more resolution
+
+if (nargin < 3)
+   bins_multiplier = 2;
+end
+
+% Flip vector if it's a column vector
+if (size(y,2) == 1)
+   y = y';
+end
+
+% Note: I'm leaving out high frequencies by choosing highest_key = 4186.01, so we might want to raise this to improve classification
+
+% Compute the sparse kernel for the fast constant-q transform
+semitones_per_octave = 12;
+lowest_key = 27.5;
+highest_key = 4186.01;
+sparkernel = sparse_kernel(lowest_key, highest_key, semitones_per_octave*bins_multiplier, fs);
+
+% Initialize the Q-gram matrix
+spec_bins = size(sparkernel,2);
+step_size = .01*fs; % 10 ms
+steps = (length(y) - size(sparkernel,1))/step_size + 1;
+Q = zeros(spec_bins, steps);
+
+% Compute the constant-q transform at every .01*fs seconds
+for i=1:steps
+   q = const_q_fast(y(1+(i-1)*step_size:end), sparkernel);
+   Q(:,i) = q; 
+end
+
+% To plot:
+% imagesc(abs(Q));
+
+end

### 2013-12-04 Music/timmy.sh
Removed the redundant Music folder. All midi and wav files should go in data.

@@ -1,17 +1,0 @@
-#!/bin/bash
-# Syntax:
-# ./timmy song.mid
-# outputs song.wav (keeps the original name)
-
-# Note on the 16k sampling rate:
-# The SVM paper uses an 8k sampling rate, but
-# I noticed a significant quality drop at 8k, 
-# whereas 16k seems to preserve the original quality
-# Note, too, that the highest key on the piano is 4186Hz,
-# which means 8k is insufficient, but this note also contains 
-# harmonics, so I think 16k is a good sampling rate to capture
-# these overtones.
-
-output=${1%.*}".wav"
-timidity $1 -OwM -o $output -s 16k
-

### 2013-12-05 bin/poliner_svm.m
more error stuff

@@ -31,6 +31,7 @@ function [estimated_pr, pr] = poliner_svm(wav_file, midi_file_1, midi_file_2)
     [S, f, spec_t] = specgram(y, 2048, fs, hanning(.128*fs), 2048 - 160); % Needs to be spectrogram in MATLAB  
   else
     [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
+    %S = qgram(y, fs); 
   end
 
   size(S) 
@@ -260,8 +261,14 @@ function [estimated_pr, pr] = poliner_svm(wav_file, midi_file_1, midi_file_2)
   TP_S=0;
   FP_S=0;  
   FN_S=0;
-  ETOT_S=0;
   ETOT_R=0;
+  ESUB_R=0;
+  EMISS_R=0;
+  EFA_R=0;
+  ETOT_S=0;
+  ESUB_S=0;
+  EMISS_S=0;
+  EFA_S=0;
   for j =1:subset
     for i =1:size(pr,1)
       
@@ -287,18 +294,32 @@ function [estimated_pr, pr] = poliner_svm(wav_file, midi_file_1, midi_file_2)
       
       %2nd error measure
       N_ref = sum(pr(:,j)==1);
-      N_sys_S = sum(smooth_labels(:,j)==1);
-      N_sys_R = sum(estimated_pr(:,j)==1);
-      N_corr_S =   sum((pr(:,j)==1)+ (smooth_labels(:,j)==1)==2);
-      N_corr_R =   sum((pr(:,j)==1)+ (estimated_pr(:,j)==1)==2);
       
-      ETOT_S = ETOT_S+max(N_ref,N_sys_S)-N_corr_S;
-      ETOT_R = ETOT_R+max(N_ref,N_sys_R)-N_corr_R;
+      N_sys_S = sum(smooth_labels(:,j)==1);
+      N_corr_S =   sum((pr(:,j)==1)+ (smooth_labels(:,j)==1)==2);
+      ETOT_S = ETOT_S+max(N_ref,N_sys_S)-N_corr_S; 
+      ESUB_S = ESUB_S+min(N_ref,N_sys_S)-N_corr_S;
+      EMISS_S = EMISS_S+max(0,(N_ref-N_sys_S));
+      EFA_S = EFA_S+max(0,(N_sys_S-N_ref));
+      
[…]

### 2013-12-05 bin/qgram.m
Added f and spec_t, like spectrogram, to qgram

@@ -1,4 +1,4 @@
-function Q = qgram(y, fs, bins_multiplier)
+function [Q, f, spec_t] = qgram(y, fs, bins_multiplier)
 % y is the signal
 % fs is the sampling frequency
 % bins_multiplier multiples the number of bins per octave. When bins_multiplier=1, bins=12 (which is the number of semitones per octave). It seems like researchers typically choose bins_multiplier=2 for more resolution
@@ -30,6 +30,18 @@ Q = zeros(spec_bins, steps);
 for i=1:steps
    q = const_q_fast(y(1+(i-1)*step_size:end), sparkernel);
    Q(:,i) = q; 
+end
+
+% Compute the frequency grid
+f = zeros(1,spec_bins);
+for i=1:spec_bins
+    f(i) = lowest_key * 2^((i-1)/semitones_per_octave/bins_multiplier);
+end
+
+% Compute the time grid
+spec_t = zeros(1, steps);
+for i=1:steps
+   spec_t(i) = .01*(i-1);
 end
 
 % To plot:

### 2013-12-05 bin/calc_error.m
moved error stuff into calc_error

@@ -1,0 +1,43 @@
+function [Acc, E_tot, E_sub, E_miss, E_fa] = calc_error( true_pr, observed_pr, subset )
+%UNTITLED Summary of this function goes here
+%   Detailed explanation goes here
+  TP=0;
+  FP=0;
+  FN=0;
+  ETOT=0;
+  ESUB=0;
+  EMISS=0;
+  EFA=0;
+  for j =1:subset
+    for i =1:size(true_pr,1)
+      
+          tl = true_pr(i,j);
+          ol = observed_pr(i,j);
+          
+          if (tl==1 && ol==1)
+              TP=TP+1;              
+          elseif (tl==0 && ol ==1)
+              FP=FP+1;
+          elseif (tl==1 && ol ==0)
+              FN=FN+1;
+          end
+    end      
+    %2nd error measure
+    N_ref = sum(true_pr(:,j)==1);
+      
+    N_sys = sum(observed_pr(:,j)==1);
+    N_corr =   sum((true_pr(:,j)==1)+ (observed_pr(:,j)==1)==2);
+    ETOT = ETOT+max(N_ref,N_sys)-N_corr; 
+    ESUB = ESUB+min(N_ref,N_sys)-N_corr;
+    EMISS = EMISS+max(0,(N_ref-N_sys));
+    EFA = EFA+max(0,(N_sys-N_ref));
+  end
+  Norm = sum(sum(true_pr(:,1:subset)==1));
+  Acc=(TP)/(TP+FP+FN); %0.4689 for 50, .54 for 100
+  E_tot = ETOT/Norm; % number of frames that are actually on
+  E_sub = ESUB/Norm;
+  E_miss = EMISS/Norm;
+  E_fa = EFA/Norm;
+
+end
+

### 2013-12-05 bin/poliner_svm.m

@@ -31,11 +31,11 @@ function [estimated_pr, pr] = poliner_svm(wav_file, midi_file_1, midi_file_2)
     [S, f, spec_t] = specgram(y, 2048, fs, hanning(.128*fs), 2048 - 160); % Needs to be spectrogram in MATLAB  
   else
     [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
-    %S = qgram(y, fs); 
+    %[S, f, spec_t] = qgram(y, fs,1); 
   end
 
   size(S) 
-  spec_t(end)
+  %spec_t(end)
   % pause
   % Compute the FFT magnitudes
   magS = abs(S);
@@ -93,10 +93,10 @@ function [estimated_pr, pr] = poliner_svm(wav_file, midi_file_1, midi_file_2)
       pos = find(note_vec == 1);
 
       num_examples = 0;
-      if length(pos) < 100
+      if length(pos) < 200
       	 num_examples = length(pos);
       else
-	 num_examples = 100;
+	 num_examples = 200;
          pos = pos(randperm(length(pos)));
       end
 
@@ -110,16 +110,17 @@ function [estimated_pr, pr] = poliner_svm(wav_file, midi_file_1, midi_file_2)
       neg_examples = magS(:,neg);
       
       midi_note = nn(i);
-      if (midi_note >= 21 && midi_note <= 83) %0-2K
-          pos_examples = pos_examples(1:256, :);
-          neg_examples = neg_examples(1:256, :);
-      elseif (midi_note > 83 && midi_note <= 95) %1K-3K
-          pos_examples = pos_examples(129:384, :);
-          neg_examples = neg_examples(129:384, :);              
-      else                                          %2K-4K
-          pos_examples = pos_examples(257:512, :);
-          neg_examples = neg_examples(257:512, :);   
-      end
+      
+%       if (midi_note >= 21 && midi_note <= 83) %0-2K
+%           pos_examples = pos_examples(1:256, :);
+%           neg_examples = neg_examples(1:256, :);
+%       elseif (midi_note > 83 && midi_note <= 95) %1K-3K
+%           pos_examples = pos_examples(129:384, :);
+%           neg_examples = neg_examples(129:384, :);              
[…]

### 2013-12-06 bin/nmf_transcribe.m
Added WIP nmf transcribe script

@@ -1,0 +1,75 @@
+  % Warning: this is written for Octave and not tested with MATLAB
+  % This also uses libsvm. The libsvm I have included has been built for Octave
+
+  fs = 0;
+  filelist = readdir('../data/notes');
+  for i = 1:numel(filelist)
+    if (regexp (filelist{i},'^\w+.wav$'))
+        wavs{i} = filelist{i};
+    end
+  end
+
+  W = [];
+  for i = 1:numel(wavs)
+    if (~isempty(wavs{i}))
+      wavs{i} = strcat('../data/notes/', wavs{i});
+      disp(wavs{i});
+      [y,fs,bps] = wavread(wavs{i});
+      [S, f, spec_t] = qgram(y, fs,1); 
+
+      Sen = abs(sum(S));
+      S(:, Sen < 0.00001) = zeros(88,sum(Sen < 0.00001));
+
+      Sm = abs(S(:,30))';
+      Sm2 = sm;
+      g = mean(S(:, j));
+      Sm2 = log(1 + (1/g).*
+
+      plot(Sm);
+      drawnow;
+      W = [W; Sm];
+    end
+  end
+
+  imagesc(W);
+
+    pause;
+  num_bins = 176;
+% 
+% for i = 1:numel(wavs)
+%   if (wavs{i} = [])
+%   disp(wavs{i});
+% end
+
+    pause;
+
+  more off;
+  wav_file = '../data/br_im2.wav';
+  % Open wav file
+  [y,fs,bps] = wavread(wav_file);
+  % length(y)
+  % fs
+  % .128*fs
+  %.01*fs
+  % Compute the STFT on 128ms frames with 10ms hops
+  isOctave = exist('OCTAVE_VERSION') ~= 0;
+  if isOctave
+    [S, f, spec_t] = specgram(y, 2048, fs, hanning(.128*fs), 2048 - 160); % Needs to be spectrogram in MATLAB  
+  else
+    [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
+  end
+
+  size(S) 
+  spec_t(end)
+  % pause
+  % Compute the FFT magnitudes
+  magS = abs(S);
+
+  % Normalize, whiten the FFT magnitudes (to-do)
+    
+  % Add the libsvm and midi libraries  
+  addpath('../lib/libsvm');
+  addpath('../lib/nmf');
+
+  [W, H, iter, ~] = nmf(magS, 88, 'verbose',2, 'type', 'sparse');
+

### 2013-12-06 bin/qgram.m
Steps should be rounded before being used as a size

@@ -24,7 +24,7 @@ sparkernel = sparse_kernel(lowest_key, highest_key, semitones_per_octave*bins_mu
 spec_bins = size(sparkernel,2);
 step_size = .01*fs; % 10 ms
 steps = (length(y) - size(sparkernel,1))/step_size + 1;
-Q = zeros(spec_bins, steps);
+Q = zeros(spec_bins, floor(steps));
 
 % Compute the constant-q transform at every .01*fs seconds
 for i=1:steps
@@ -39,7 +39,7 @@ for i=1:spec_bins
 end
 
 % Compute the time grid
-spec_t = zeros(1, steps);
+spec_t = zeros(1, floor(steps));
 for i=1:steps
    spec_t(i) = .01*(i-1);
 end

### 2013-12-06 bin/nmf_transcribe.m
Made file loading octave and matlab compatable

@@ -2,11 +2,23 @@
   % This also uses libsvm. The libsvm I have included has been built for Octave
 
   fs = 0;
-  filelist = readdir('../data/notes');
-  for i = 1:numel(filelist)
-    if (regexp (filelist{i},'^\w+.wav$'))
-        wavs{i} = filelist{i};
-    end
+  
+  % Get list of training notes
+  isOctave = exist('OCTAVE_VERSION') ~= 0;
+  if isOctave
+         filelist = readdir('../data/notes');
+      for i = 1:numel(filelist)
+        if (regexp (filelist{i},'^\w+.wav$'))
+            wavs{i} = filelist{i};
+        end
+      end
+  else
+      filelist = dir('../data/notes');
+      for i = 1:numel(filelist)
+        if (regexp (filelist(i).name,'^\w+.wav$'))
+            wavs{i} = filelist(i).name;
+        end
+      end
   end
 
   W = [];
@@ -18,12 +30,12 @@
       [S, f, spec_t] = qgram(y, fs,1); 
 
       Sen = abs(sum(S));
-      S(:, Sen < 0.00001) = zeros(88,sum(Sen < 0.00001));
+      S(:, Sen <  0.0000001) = zeros(88,sum(Sen < 0.0000001));
 
       Sm = abs(S(:,30))';
-      Sm2 = sm;
-      g = mean(S(:, j));
-      Sm2 = log(1 + (1/g).*
+      %Sm2 = sm;
+      %g = mean(S(:, j));
+      %Sm2 = log(1 + (1/g).*
 
       plot(Sm);
       drawnow;

### 2013-12-06 bin/nmf_transcribe.m
Finished making the dictionary matrix.
The script will use the dictionary as initial value for W. I still need to
add the modified, projected gradient descent that only updates H.

@@ -24,64 +24,47 @@
   W = [];
   for i = 1:numel(wavs)
     if (~isempty(wavs{i}))
+        
+      % Read File
       wavs{i} = strcat('../data/notes/', wavs{i});
       disp(wavs{i});
       [y,fs,bps] = wavread(wavs{i});
+      
+      % Constant Q Transform
       [S, f, spec_t] = qgram(y, fs,1); 
 
-      Sen = abs(sum(S));
-      S(:, Sen <  0.0000001) = zeros(88,sum(Sen < 0.0000001));
+      % Zero-out spectra with low energy (for noise reduction)
+      Sen = sum(abs(S).^2);
+      S(:, Sen <  1e-9) = zeros(88,sum(Sen < 1e-9));
 
-      Sm = abs(S(:,30))';
-      %Sm2 = sm;
-      %g = mean(S(:, j));
-      %Sm2 = log(1 + (1/g).*
-
-      plot(Sm);
+      % Average the signal-containing spectra
+      Sm = mean(abs(S)');
+     
+      % Magnitude warping and mean subtraction proposed by Neidermeyer
+      g = mean(Sm); % What average should we use (mean? outlier resistant?)
+      S_warp = log(1 + (1/g).*Sm);
+      y = smooth(S_warp, 9);
+      wi = max(0, S_warp - y');
+      
+      plot(wi);
       drawnow;
-      W = [W; Sm];
+      W = [W; wi];
     end
   end
-
+  W = W./(max(max(W)));
   imagesc(W);
 
-    pause;
-  num_bins = 176;
-% 
-% for i = 1:numel(wavs)
-%   if (wavs{i} = [])
-%   disp(wavs{i});
-% end
-
-    pause;
-
-  more off;
+%%
   wav_file = '../data/br_im2.wav';
   % Open wav file
   [y,fs,bps] = wavread(wav_file);
-  % length(y)
-  % fs
-  % .128*fs
-  %.01*fs
-  % Compute the STFT on 128ms frames with 10ms hops
-  isOctave = exist('OCTAVE_VERSION') ~= 0;
-  if isOctave
-    [S, f, spec_t] = specgram(y, 2048, fs, hanning(.128*fs), 2048 - 160); % Needs to be spectrogram in MATLAB  
-  else
-    [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
-  end
 
-  size(S) 
[…]

### 2013-12-08 bin/nmf_transcribe.m
Wrote non-negative projected gradient descent for NMF
I've tried basic linear STFT and Q-Gram for features, neither work that well.

@@ -31,7 +31,8 @@
       [y,fs,bps] = wavread(wavs{i});
       
       % Constant Q Transform
-      [S, f, spec_t] = qgram(y, fs,1); 
+      %[S, f, spec_t] = qgram(y, fs,1); 
+      [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
 
       % Zero-out spectra with low energy (for noise reduction)
       Sen = sum(abs(S).^2);
@@ -40,7 +41,7 @@
       % Average the signal-containing spectra
       Sm = mean(abs(S)');
      
-      % Magnitude warping and mean subtraction proposed by Neidermeyer
+      % Magnitude warping and mean subtraction proposed by Niedermeyer
       g = mean(Sm); % What average should we use (mean? outlier resistant?)
       S_warp = log(1 + (1/g).*Sm);
       y = smooth(S_warp, 9);
@@ -59,12 +60,81 @@
   % Open wav file
   [y,fs,bps] = wavread(wav_file);
 
-  [S, f, spec_t] = qgram(y, fs,1); 
+  %[S, f, spec_t] = qgram(y, fs,1);
+   [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
 
-    
+ %% 
   % Add the libsvm and midi libraries  
   addpath('../lib/libsvm');
   addpath('../lib/nmf');
-pause;
-  [W2, H, iter, ~] = nmf(abs(S), 88, 'verbose',2, 'type', 'sparse', 'w_init', W);
+ 
+  magS = abs(S).^2;
+  magS = magS ./ (max(max(magS)));
 
+  %[W2, H, iter, ~] = nmf(abs(S), 88, 'verbose',2, 'w_init', W);
+
+  %H = rand(88, length(S));
+  %Wt = W';
+  bins = size(S,1);
+  H = [];
+for i = 1:length(magS)
+    fprintf('Testing frame: %i\n', i);
+    v = magS(:, i);
+    h = zeros(88, 1);
+    P = [];
+    Z = (1:88)';
+    
+    gradcost = Wt'*(v - Wt*h); % least squares gradient
+    skip_grad = 0;
+    iters = 0;
+    while (~isempty(Z) || any(gradcost(Z) >= 0))
+        %fprintf('Norm error %g\n', .5*norm(Wt*h - v)^2);
+        %pause
+        iters = iters + 1;
+        if (iters > 10)
+            break;
+        end
+        if (skip_grad == 0)
+
[…]

### 2013-12-08 bin/nmf_transcribe.m
ditto?

@@ -31,8 +31,8 @@
       [y,fs,bps] = wavread(wavs{i});
       
       % Constant Q Transform
-      %[S, f, spec_t] = qgram(y, fs,1); 
-      [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
+      [S, f, spec_t] = qgram(y, fs,1); 
+      %[S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
 
       % Zero-out spectra with low energy (for noise reduction)
       Sen = sum(abs(S).^2);
@@ -60,8 +60,8 @@
   % Open wav file
   [y,fs,bps] = wavread(wav_file);
 
-  %[S, f, spec_t] = qgram(y, fs,1);
-   [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
+  [S, f, spec_t] = qgram(y, fs,1);
+  %[S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
 
  %% 
   % Add the libsvm and midi libraries  
@@ -74,7 +74,7 @@
   %[W2, H, iter, ~] = nmf(abs(S), 88, 'verbose',2, 'w_init', W);
 
   %H = rand(88, length(S));
-  %Wt = W';
+  Wt = W';
   bins = size(S,1);
   H = [];
 for i = 1:length(magS)

### 2013-12-09 paper/paper.tex
Added comments about pictures to the paper
added section about error evaluation

@@
 \address[csadd]{Computer Science Department, University of Michigan, Ann Arbor, MI 48109}
+
+% Pictures to add to paper:
+% spectrogram of single note
+% spectrogram of multiple notes
+
+% Show FFT vs. Constant q or log bins
+
+% Show SVM output
+% Show SVM with HMM output
+
+% Show NMF output and NMF output with sparsity constraint
+% Show NMF with Itakura-Saito Divergence
 
@@
 
-Polyphony means multiple simultaneous notes. Monophony means only a single note at a time. These terms carry slightly different meanings outside of the automated transcription field.
+Polyphony means multiple simultaneous notes. Monophony means only a single note at a time. These terms carry slightly different meanings outside of automated transcription research.
 
@@
 
-\subsubsection{}
+We use a few metrics for evaluation, as per the existing literature.
 
-\subsubsection{Frame-level transcription error score}
+\subsubsection{Overall accuracy}
+$$Acc = \frac{TP}{TP+FP+FN} $$
+where TP (true positives) is the number of correctly transcribed notes (over all notes), FP (false positives) is the number of unvoiced frames transcribed as voiced, and FN (false negatives) is the number of voiced frames transcribed as unvoiced.
+Note that notes that are missed and notes that are inserted are weighed equally.
+\subsection{Frame-level transcription error score}
+In the previous measure "double counts" notes transcribed correctly but at the wrong time.  Frame-level transcription error score seeks to avoid that.
+At each time frame, let $N_{sys}$ be the number of reported pitches, $N_{ref}$ the ground truth number of pitches, and $N_{corr}$, their intersection, is the number of correctly reported pitches.
+Then we define
+$$E_{tot}=\frac{\sum_{t=1}^T \text{max}(N_{ref}(t),N_{sys}(t))-N_{corr}(t)}{\sum_{t=1}^T N_{ref}(t)}$$
+$E_{tot}$ can be further split into three separate scores: $E_{sub}$, which counts the number (at each frame) of ground truth notes for which some other note was reported, $E_{miss}$, the number of ground truth notes which cannot be accounted for, and $E_{fa}$, the number of reported notes which cannot be paired with a ground truth note.
+The equations for these, respectively, are
+$$E_{sub}=\frac{\sum_{t=1}^T \text{min}(N_{ref}(t),N_{sys}(t))-N_{corr}(t)}{\sum_{t=1}^T N_{ref}(t)}$$
+$$E_{miss}=\frac{\sum_{t=1}^T \text{max}(0,N_{ref}(t))-N_{sys}(t)}{\sum_{t=1}^T N_{ref}(t)}$$
+$$E_{fa}=\frac{\sum_{t=1}^T \text{max}(0,N_{sys}(t))-N_{ref}(t)}{\sum_{t=1}^T N_{ref}(t)}$$
 
+\subsubsection{Note Onset Detection}
+Frame-level transcription error requires a high level of precision.
+An error-measure perhaps more fitting for this problem is note onset detection, since onset is usually more important than offset.
+To be counted as correct, our transcribed note must be turned "on" within $100$ milliseconds of the ground truth onset.
 \subsection{Method comparison}

### 2013-12-09 paper/paper.tex
added hmm section

@@
 \subsubsection{HMM smoothing}
+In many of the techniques used (such as SVM's), frames are classified independently of their neighbors.
+Since this does not take advantage of the temporal structure of piano music, Hidden Markov Models (HMM's) are often used for post-processing.
+We model each note independently using a two-state (corresponding to on/off) HMM.
+The transition matrix and state priors can be easily estimated from the ground truth.
+The emission matrix can also be estimated by applying the main classifier to the training data.
+With these estimates, we can smooth the output by finding the most likely state sequence using the Viterbi algorithm.
 \subsubsection{Probabilistic spectral smoothness}

### 2013-12-10 bin/nmf_transcribe.m
Worked on it

@@ -1,6 +1,6 @@
   % Warning: this is written for Octave and not tested with MATLAB
   % This also uses libsvm. The libsvm I have included has been built for Octave
-
+  more off;
   fs = 0;
   
   % Get list of training notes
@@ -44,16 +44,22 @@
       % Magnitude warping and mean subtraction proposed by Niedermeyer
       g = mean(Sm); % What average should we use (mean? outlier resistant?)
       S_warp = log(1 + (1/g).*Sm);
-      y = smooth(S_warp, 9);
-      wi = max(0, S_warp - y');
       
-      plot(wi);
-      drawnow;
-      W = [W; wi];
+      %%y = smooth(S_warp, 9);
+      y = filter(ones(9,1)/9, 1, S_warp);
+      wi = max(0, S_warp - y);
+      
+      %plot(wi);
+      %drawnow;
+      W = [W wi'];
     end
   end
-  W = W./(max(max(W)));
+  W = abs(W);
+  for i = 1:88
+    W(:, i)  = W(:, i) ./ norm(W(:, i), 2); 
+  end
   imagesc(W);
+  pause;
 
 %%
   wav_file = '../data/br_im2.wav';
@@ -66,9 +72,9 @@
  %% 
   % Add the libsvm and midi libraries  
   addpath('../lib/libsvm');
-  addpath('../lib/nmf');
+  addpath('../lib/matlab-midi/src');
  
-  magS = abs(S).^2;
+  magS = abs(S);
   magS = magS ./ (max(max(magS)));
 
   %[W2, H, iter, ~] = nmf(abs(S), 88, 'verbose',2, 'w_init', W);
@@ -77,64 +83,82 @@
   Wt = W';
   bins = size(S,1);
   H = [];
-for i = 1:length(magS)
-    fprintf('Testing frame: %i\n', i);
-    v = magS(:, i);
-    h = zeros(88, 1);
-    P = [];
-    Z = (1:88)';
-    
-    gradcost = Wt'*(v - Wt*h); % least squares gradient
-    skip_grad = 0;
-    iters = 0;
-    while (~isempty(Z) || any(gradcost(Z) >= 0))
-        %fprintf('Norm error %g\n', .5*norm(Wt*h - v)^2);
-        %pause
-        iters = iters + 1;
-        if (iters > 10)
-            break;
-        end
-        if (skip_grad == 0)
+ H = rand(88, 2000);
+ V = magS(:, 1:2000);
[…]

### 2013-12-10 bin/calc_error.m
Added key for what those variables mean

@@ -1,6 +1,14 @@
 function [Acc, E_tot, E_sub, E_miss, E_fa] = calc_error( true_pr, observed_pr, subset )
 %UNTITLED Summary of this function goes here
 %   Detailed explanation goes here
+
+% Acc - Overall Accuracy
+% E_tot  - total fram level error
+% E_sub  - Substitution (Some note was reported but not the correct one)
+% E_miss - Missed 
+% E_fa   - False Alarms
+
+
   TP=0;
   FP=0;
   FN=0;

### 2013-12-10 bin/calc_error.m
Added two more error metrics

@@ -1,4 +1,4 @@
-function [Acc, E_tot, E_sub, E_miss, E_fa] = calc_error( true_pr, observed_pr, subset )
+function [Acc, E_tot, E_sub, E_miss, E_fa, precision, recall, f] = calc_error( true_pr, observed_pr, subset )
 %UNTITLED Summary of this function goes here
 %   Detailed explanation goes here
 
@@ -46,6 +46,10 @@ function [Acc, E_tot, E_sub, E_miss, E_fa] = calc_error( true_pr, observed_pr, s
   E_sub = ESUB/Norm;
   E_miss = EMISS/Norm;
   E_fa = EFA/Norm;
+  
+  precision = (TP) / (TP  + FP);
+  recall = (TP) / (TP  + FN);
+  f = 2*precision*recall / (precision + recall);
 
 end
 

### 2013-12-10 paper/paper.tex
Added adaboost algorithm, fixed octave offset in qgram, made a qgram cache function, added adaboost libraries, and added sections to the paper

--- 2013-12-09 paper/paper.tex
@@
 \documentclass[5p]{elsarticle}
+\usepackage{epstopdf}
+\epstopdfsetup{outdir=./}
+\DeclareGraphicsRule{.eps}{pdf}{.pdf}{`epstopdf #1}
 \usepackage{natbib,graphicx,amsmath, mhchem,url,colortbl,fancyhdr,subfigure}
@@
 
+In Western music, each note corresponds to a single, fundamental frequency.
+The musical note A above middle C corresponds to 440Hz.
+If this were the complete story, there would be a one-to-one correspondence between frequencies and notes, and the transcription problem would be as simple as converting the frequencies into notes.
+The reality, however, is that each instrument has a unique timbre with a set of overtones that vary over time and overlap with the frequencies of other notes.
+
+Like most research in automatic music transcription, we restrict our problem to transcribing polyphonic piano music.
+The reason for this is primarily the availability of online piano MIDI databases and the popularity of the piano in Western music.
+At the same time, the piano has a complex timbre and adequately represents the difficulties in the generic transcription problem.
+
+\subsection{Piano physics}
+
+\citet{bank2003physically} details the physics of piano sound synthesis. We will just note here that...
+
+Figure X shows the spectral profile of a low C note played over a couple seconds.
+
+This is typically the lowest 
+(But it's more complicated than that...)
+- Show why it's complicated 
+- Show some of the physics
+
 \section{Related Work}
@@
 
+Our training data comes from a large database of piano MIDI files. MIDI is a standard for specifying...
+%musical score
+%While MIDI specifies the type of instrument, 
+% MIDI instrument not specified in the file 
+%a standard for specifying
+
+%Wav files are...
+
+%44.1kHz
+
+%16kHz sampling rate, 16 bit,
+%Databases
+
 % Sampling rate for piano
@@
 % Hint at the matrix factorization approach
+
+In all of the methods we explored, we computed a spectrogram of the wav file with a window size of 128ms, a Hanning windowing function, and a 10ms hop between frames.
+This window size gives us a frequency resolution of 8Hz, which is slightly inadequate for resolving the fundamental frequencies of lower notes and overly resolute for the higher frequencies due to the logarithmic spacing of the semitone fundamental frequencies.
+
+To the authors' knowledge, all of the methods in the transcription literature compute a spectrogram as a preprocessing step before applying their specialized algorithm.
+In the NMF methods, the spectrogram matrix is decomposed into a piano roll matrix and the note spectrum dictionary matrix.
+In the supervised classifier methods, the classifiers use features from the spectrogram matrix to predict the presence of a note.
+
+\begin{figure*}[t]
+\begin{center}
+\subfigure[]{
+\includegraphics[scale=.6]{pianorollsong}
+\label{fig:pianorollsong}
+}
+\subfigure[]{
+\includegraphics[scale=.6]{brim2ud}
+\label{fig:brim2ud}
+}
+\end{center}
+\caption{(a) shows the piano roll generated from a MIDI file.
+(b) shows a spectrogram of the wav file using the constant Q transform with a step size of 10ms.
+One can immediately see the similarity in the shape of the two plots, although the transcription process is complicated by the overlapping frequencies, changing volume, and noise in the spectrogram plot.}
+\end{figure*}
+
 \subsection{Constant Q transform}
-
-\subsection{Semitone filter bank}
 Unlike the linear frequency scale used in an FFT, semitones in Western music are spaced logarithmically.
@@
 It reduces the dimensionality of the training data, since most of the spectral energy occupies a logarithmic grid of frequencies.
-In practice, \citet{bock2012polyphonic} were able to reduce the dimensionality of their training data from 5120 FFT magnitudes down to 183.
+In practice, \citet{bock2012polyphonic} were able to reduce the dimensionality of their training data from 2048 FFT magnitudes down to 183.
 
@@
 Most researchers in this field evaluate their music transcription methods on their note onset detection performance.
+
 \subsection{Discriminative models}
@@
 \subsection{SVM}
-% Frequency partition
+
+% one vs all
+\citet{poliner2006discriminative}
+
+% frequency partition
+
+\subsection{Restricted Boltzmann machine}
+
+\citet{nam2011classification} proposes ...
+ training a restricted Boltzmann machine on the spectrogram features. The features computed by the RBM are used by an SVM
+
+\subsection{Boosted classifier}
+
+\citet{boogaart2009note} proposes a cascade of boosted classifiers.
 
@@
 
-The KL divergence cost function is more often used by transcription researchers because it XXX.
+% \citet{lee1999learning}
+The functions $||V-WH||^{2}$ and $D(V||WH)$ are convex in W only and H only, but they are not convex in W and H together.
+There is no known procedure for finding the global minimum of these functions, but \citet{seung2001algorithms} have proposed methods for updating W and H such that the measures above will decrease or remain constant after every update.
+Their paper proposes a multiplicative update rule and a gradient descent rule.
 
-\citet{lee1999learning}
-\citet{seung2001algorithms}
+The multiplicative update rule for improving the Euclidean distance error measure $||V-WH||^{2}$ is:
+
+\[ H_{\alpha \mu} \leftarrow H_{\alpha \mu} \frac{ (W^{T} V)_{\alpha \mu}}{(W^{T} W H )_{\alpha \mu}}, W_{\gamma \alpha} \leftarrow W_{\gamma \alpha} \frac{ (V H^{T})_{\gamma \alpha}} {(W H H^{T})_{\gamma \alpha}} \]
+
+\noindent Typically, $W$ and $H$ are initialized with random, nonnegative values and iterated until convergence.
+
+% Applying NMF to the piano transcription problem is not as straightforward as computing the above update rules.
+
+There are a few problems applying NMF as it is presented above to the piano transcription problem:
+
+\begin{description}
+\item[$\bullet$] The algorithms from \citet{seung2001algorithms} require that the number of keys present in the song is specified in advance.
+While a full piano keyboard contains a constant 88 keys, a piano song will often only use a subset of all the keys.
+In practice, pre-selecting the $W$ matrix to find more keys than are present results in the algorithm discovering parts of notes or the same note at different points in the note's ADSR envelope; for example, the attack of a note can have a different spectral signature than its release.
+\citet{weninger2013discriminative} use this fact explicitly to learn notes at different points in the ADSR envelope using NMF as a preprocessing step to SVM classification.
+\citet{} uses cross-validation to select the appropriate number of keys.
+\item[$\bullet$] The piano roll matrix $H$ and the spectrogram $V$ are large and often dense matrices, so convergence using the multiplicative update rules can take on the order of minutes on a modern laptop.
+\item[$\bullet$] The notes represented in the $W$ and $H$ matrix are not ordered properly. \citet{} resorted to ordering the notes in $W$ by hand. 
+\end{description}
 
@@
 This is desirable in music transcription because the piano roll matrix, $H$, is a sparse matrix of note activations.
-However, the NMF method does not contain an explicit sparsity objective.
+However, the NMF method does not contain an explicit sparsity objective, and so the update rules do not always converge to a sparse solution.
 \citet{hoyer2004non} extended NMF to contain a tunable sparsity objective.
-
-The sparse NMF method in \citet{hoyer2004non} was adopted by \citet{abdallah2004polyphonic} in their transcription system.
+This sparse NMF method in \citet{hoyer2004non} was adopted by \citet{abdallah2004polyphonic} in their transcription system.
 
 % Non-negative Matrix factorization
+
+\subsection{Supervised NMF}
+
+In the previous NMF methods, we extracted both the note dictionary matrix and the piano roll from the spectrogram.
+In piano transcription, however, the dictionary matrix should be similar across multiple transcriptions.
+We can use
+
 \subsection{Bayesian nonparametric models}
+One of the downsides of the NMF methods is that they require the number of notes (or sources) to be specified
+
 \subsection{Smoothing}
 \subsubsection{HMM smoothing}
+In many of the techniques used (such as SVM's), notes at a single time step are classified independently of the notes at adjacent time steps.
+Since these methods do not take advantage of the temporal structure of piano music, hidden Markov models are often used for post-processing the output of the discriminative models.
+We model each note independently using a two-state HMM, where the states correspond to the note being on or off.
+The transition matrix and state priors can be estimated from the ground truth piano roll, while the emission matrix can be estimated using the output from the classifier as the observable variables.
+Applying the Viterbi algorithm to each note class results in a smoother piano roll that eliminates unlikely, spurious, isolated notes.
+
 \subsubsection{Probabilistic spectral smoothness}
+
 \subsection{Recurrent neural network}
@@
 
-There is no standard metric to evaluate the performance of a transcription method. However, many researchers use the following methods:
+\subsection{MIDI databases}
 
-\subsection{MIDI databases}
+The MIDI music used to evaluate these methods came from the Classical Piano MIDI Page at http://www.piano-midi.de/, a free, online database of piano MIDI files with a wide range of composers.
+This database, along with the MAPS database, are a popular choice for piano transcription researchers.
+Table X gives a list of the composers and songs we used for testing.
+
+The MIDI files represent the ground truth data. To generate training data, we synthesized wav files from the MIDI files using Timidity...
+
+Synthesized with Timidity -> 16kHz
 
@@
 
-We use a few metrics for evaluation, as per the existing literature.
+% We use a few metrics for evaluation, as per the existing literature.
+
+There is no standard metric to evaluate the performance of a transcription method. However, many researchers use the following methods:
 
@@
 Sparse NMF & 24\% & 24 \\ \hline
+AdaBoost & 25\% & 25 \\ \hline
 \end{tabular}
@@
 Sparse NMF & 21\% & 21 \\ \hline
+AdaBoost & 22\% & 22 \\ \hline
 \end{tabular}

### 2013-12-10 bin/estimate_hmm.m

@@ -1,0 +1,61 @@
+function [priors_cell, trans_mat_cell, em_mat_cell] = estimate_hmm(raw_pr, nn, true_pr)
+  %%% Estimate HMM %%%
+  % Run a hidden markov model over the piano roll for smoothing the raw log posterior probabilities
+  % Estimation of priors and transition matrix can be done using MIDI files (ground truth)
+  % Estimation of emission matrix should be done using predict_label from SVM
+  
+  % 1 is off
+  % 2 is on
+
+  num_frames = size(raw_pr, 2);
+  num_notes = size(nn,2);
+
+  trans_mat_cell = cell(1,num_notes);
+  em_mat_cell = cell(1,num_notes);
+  priors_cell = cell(1,num_notes);
+
+  % Estimate a HMM for each note
+  for i=1:size(nn,2)
+    fprintf('Estimating HMM %d\n', i);
+    cur_note = nn(i);
+
+    % If cur_note is not actually played in song, skip it
+    %if (isempty(find(notes_list == cur_note, 1)))
+    %    continue;
+    %end
+
+    % Estimate prior
+    prior_on = sum(true_pr(i,:)) / num_frames;
+    prior_off = 1 - prior_on;
+
+    % Estimate transition matrix
+    trans_mat = zeros(2,2);
+    prev_frame = 0;
+    for j=1:num_frames
+    	cur_frame = true_pr(i,j);
+	trans_mat(prev_frame+1, cur_frame+1) = trans_mat(prev_frame+1, cur_frame+1) + 1;
+	prev_frame = cur_frame;
+    end
+    trans_mat = trans_mat ./ num_frames;
+
+    % Estimate emission matrix
+    em_mat = zeros(2,2);
+    for j=1:num_frames
+    	% 1 gr, 1 emission
+
+	em_mat(true_pr(i,j)+1, raw_pr(i,j)+1) = em_mat(true_pr(i,j)+1, raw_pr(i,j)+1) + 1; 
+    end
+    em_mat =  em_mat ./ num_frames;
+
+    % Save results to cells
+    trans_mat_cell{i} = trans_mat;
+    em_mat_cell{i} = em_mat;
+    priors_cell{i} = [prior_off prior_on];
+    trans_mat
+    em_mat
+    [prior_ff prior on]
+    
+  end
+
+end
+

### 2013-12-10 bin/qgram_cache.m

@@ -1,0 +1,15 @@
+function [absQ, f, spec_t] = qgram_cache(wav_file)
+  % Note: I'm returning absQ instead of Q. We never use Q and absQ saves space
+  more off;
+  mat_cache = strcat(wav_file, '.cache');
+  try 
+    load(mat_cache);
+    fprintf('Cached qgram found!\n');
+  catch 
+    fprintf('Cached qgram does not exist. Creating new qgram\n');
+    [y,fs,g] = wavread(wav_file);
+    [Q, f, spec_t] = qgram(y,fs,2);
+    absQ = abs(Q);
+    save(mat_cache, 'absQ', 'f', 'spec_t');
+  end
+end

### 2013-12-10 lib/GML_AdaBoost_Matlab_Toolbox_0.3/Example_1.m

@@ -1,0 +1,34 @@
+% Step1: reading Data from the file
+file_data = load('Ionosphere.txt');
+Data = file_data(:,1:end-1)';
+Labels = file_data(:, end)';
+Labels = Labels*2 - 1;
+
+MaxIter = 100; % boosting iterations
+more off;
+% Step2: splitting data to training and control set
+TrainData   = Data(:,1:2:end)
+TrainLabels = Labels(1:2:end)
+pause
+
+ControlData   = Data(:,2:2:end);
+ControlLabels = Labels(2:2:end);
+
+% Step3: constructing weak learner
+weak_learner = tree_node_w(3); % pass the number of tree splits to the constructor
+
+% Step4: training with Gentle AdaBoost
+[RLearners RWeights] = RealAdaBoost(weak_learner, TrainData, TrainLabels, MaxIter);
+
+% Step5: training with Modest AdaBoost
+[MLearners MWeights] = ModestAdaBoost(weak_learner, TrainData, TrainLabels, MaxIter);
+
+% Step6: evaluating on control set
+ResultR = sign(Classify(RLearners, RWeights, ControlData));
+
+ResultM = sign(Classify(MLearners, MWeights, ControlData));
+
+% Step7: calculating error
+ErrorR  = sum(ControlLabels ~= ResultR) / length(ControlLabels)
+
+ErrorM  = sum(ControlLabels ~= ResultM) / length(ControlLabels)

### 2013-12-10 lib/GML_AdaBoost_Matlab_Toolbox_0.3/RealAdaBoost.m

@@ -1,0 +1,88 @@
+%   The algorithms implemented by Alexander Vezhnevets aka Vezhnick
+%   <a>href="mailto:vezhnick@gmail.com">vezhnick@gmail.com</a>
+%
+%   Copyright (C) 2005, Vezhnevets Alexander
+%   vezhnick@gmail.com
+%   
+%   This file is part of GML Matlab Toolbox
+%   For conditions of distribution and use, see the accompanying License.txt file.
+%
+%   RealAdaBoost Implements boosting process based on "Real AdaBoost"
+%   algorithm
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+%
+%    [Learners, Weights, final_hyp] = RealAdaBoost(WeakLrn, Data, Labels,
+%    Max_Iter, OldW, OldLrn, final_hyp)
+%    ---------------------------------------------------------------------------------
+%    Arguments:
+%           WeakLrn   - weak learner
+%           Data      - training data. Should be DxN matrix, where D is the
+%                       dimensionality of data, and N is the number of
+%                       training samples.
+%           Labels    - training labels. Should be 1xN matrix, where N is
+%                       the number of training samples.
+%           Max_Iter  - number of iterations
+%           OldW      - weights of already built commitee (used for training 
+%                       of already built commitee)
+%           OldLrn    - learnenrs of already built commitee (used for training 
+%                       of already built commitee)
+%           final_hyp - output for training data of already built commitee 
+%                       (used to speed up training of already built commitee)
+%    Return:
+%           Learners  - cell array of constructed learners 
+%           Weights   - weights of learners
+%           final_hyp - output for training data
+
+function [Learners, Weights, final_hyp] = RealAdaBoost(WeakLrn, Data, Labels, Max_Iter, OldW, OldLrn, final_hyp)
+
+if( nargin == 4)
+  Learners = {};
+  Weights = [];
+  
+  d = size(Data,2);
+  distr = ones(1, d) / d;  
+  final_hyp = zeros(1, d);
+elseif( nargin > 5)
+  Learners = OldLrn;
+  Weights = OldW;
+  if(nargin < 7)
[…]

### 2013-12-10 paper/maketex.sh

@@ -1,5 +1,5 @@
 #!/bin/bash
-pdflatex paper.tex
+pdflatex -shell-escape paper.tex
 bibtex paper
 pdflatex paper.tex
 pdflatex paper.tex

### 2013-12-10 bin/nmf_transcribe.m
It sort of works okay

@@ -21,7 +21,7 @@
       end
   end
 
-  W = [];
+  W = zeros(2049, 88);
   for i = 1:numel(wavs)
     if (~isempty(wavs{i}))
         
@@ -29,17 +29,17 @@
       wavs{i} = strcat('../data/notes/', wavs{i});
       disp(wavs{i});
       [y,fs,bps] = wavread(wavs{i});
+      y = [zeros(length(y),1); y];
       
       % Constant Q Transform
-      [S, f, spec_t] = qgram(y, fs,1); 
-      %[S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
-
-      % Zero-out spectra with low energy (for noise reduction)
-      Sen = sum(abs(S).^2);
-      S(:, Sen <  1e-9) = zeros(88,sum(Sen < 1e-9));
+      %[S, f, spec_t] = qgram(y, fs); 
+      [S, f, spec_t] = spectrogram(y, 4096, 4096-160, 4096, fs);
+      S = abs(S);
+      Sen = (sum(S).^2);
+      S(:, Sen < 100) = 0;
 
       % Average the signal-containing spectra
-      Sm = mean(abs(S)');
+      Sm = mean(S, 2);
      
       % Magnitude warping and mean subtraction proposed by Niedermeyer
       g = mean(Sm); % What average should we use (mean? outlier resistant?)
@@ -49,43 +49,54 @@
       y = filter(ones(9,1)/9, 1, S_warp);
       wi = max(0, S_warp - y);
       
-      %plot(wi);
-      %drawnow;
-      W = [W wi'];
+      plot(wi);
+      drawnow;
+      W(:, i-2) = wi';
     end
   end
-  W = abs(W);
-  for i = 1:88
-    W(:, i)  = W(:, i) ./ norm(W(:, i), 2); 
-  end
-  imagesc(W);
-  pause;
+ W = abs(W);
+ figure; imagesc(W);
+
 
 %%
   wav_file = '../data/br_im2.wav';
   % Open wav file
   [y,fs,bps] = wavread(wav_file);
 
-  [S, f, spec_t] = qgram(y, fs,1);
-  %[S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
+  %[Sy, f, spec_t] = qgram(y, fs);
+  [Sy, f, spec_t] = spectrogram(y, 4096, 4096-160, 4096, fs);
+  
[…]

### 2013-12-11 bin/midi_cache.m
Added a midi cache too, for the truly impatient

@@ -1,0 +1,15 @@
+function [pr, nn] = midi_cache(midi_file)
+  more off;
+  mat_cache = strcat(midi_file, '.cache');
+  try 
+      load(mat_cache);
+      fprintf('Cached MIDI found!\n');
+  catch		      
+      fprintf('Cached MIDI does not exist. Creating new cache.\n'); 		     
+      addpath('../lib/matlab-midi/src');
+      midi = readmidi(midi_file);
+      notes = midiInfo(midi,0);
+      [pr, g, nn] = piano_roll(notes);
+      save(mat_cache, 'pr', 'nn');
+  end
+end

### 2013-12-11 bin/adaboost_test.m
Fixed issues in the AdaBoost and my HMM code

@@ -1,0 +1,40 @@
+function adaboost_test(wav_file, midi_file)
+
+  [magS, f, spec_t] = qgram_cache(wav_file); 
+  [pr, nn] = midi_cache(midi_file);
+  load('adaboost_stumps');
+  num_notes = nn(end)-nn(1)+1;
+
+  subset = 3000;
+  view_piano_roll(spec_t(1:subset), nn, pr(:,1:subset), 'MIDI notes');
+  
+  estimated_pr = zeros(size(pr,1),subset);
+  % Predict
+  for j=1:subset
+    j
+    for i=1:num_notes
+      if sum(pr(i,:)) == 0
+      	 continue
+      end
+
+      rresult = sign(Classify(rlearners_cell{i}, rweights_cell{i}, magS(:,j)));
+      if rresult < 0
+      	 rresult = 0;
+      end
+      estimated_pr(i,j) = rresult;
+    end
+  end
+
+  view_piano_roll(spec_t(1:subset), nn, estimated_pr(:,1:subset), 'AdaBoost output');
+
+  save('adaboostout', 'estimated_pr');  
+
+  % Smooth output
+  fprintf('Smoothing output\n');
+  [priors_cell, trans_mat_cell, em_mat_cell] = estimate_hmm(estimated_pr(:,1:subset), nn, pr(:,1:subset));
+  [smooth_pr] = smooth_hmm(estimated_pr, nn, priors_cell, trans_mat_cell, em_mat_cell);   
+
+  save('smoothies', 'smooth_pr');  
+  view_piano_roll(spec_t(1:subset), nn, smooth_pr(:,1:subset), 'AdaBoost smoothed output');
+
+end

### 2013-12-11 bin/estimate_hmm.m

@@ -25,7 +25,7 @@ function [priors_cell, trans_mat_cell, em_mat_cell] = estimate_hmm(raw_pr, nn, t
     %end
 
     % Estimate prior
-    prior_on = sum(true_pr(i,:)) / num_frames;
+    prior_on = sum(true_pr(i,:)) / num_frames + 1e-5;
     prior_off = 1 - prior_on;
 
     % Estimate transition matrix
@@ -37,15 +37,22 @@ function [priors_cell, trans_mat_cell, em_mat_cell] = estimate_hmm(raw_pr, nn, t
 	prev_frame = cur_frame;
     end
     trans_mat = trans_mat ./ num_frames;
+    trans_mat(1,2) = trans_mat(1,2) / prior_off;
+    trans_mat(2,1) = trans_mat(2,1) / prior_on;
+    trans_mat(1,1) = 1 - trans_mat(1,2);
+    trans_mat(2,2) = 1 - trans_mat(2,1);
 
     % Estimate emission matrix
     em_mat = zeros(2,2);
     for j=1:num_frames
     	% 1 gr, 1 emission
-
 	em_mat(true_pr(i,j)+1, raw_pr(i,j)+1) = em_mat(true_pr(i,j)+1, raw_pr(i,j)+1) + 1; 
     end
     em_mat =  em_mat ./ num_frames;
+    em_mat(1,2) = em_mat(1,2) / prior_off;
+    em_mat(2,1) = em_mat(2,1) / prior_on;
+    em_mat(1,1) = 1 - em_mat(1,2);
+    em_mat(2,2) = 1 - em_mat(2,1);
 
     % Save results to cells
     trans_mat_cell{i} = trans_mat;
@@ -53,7 +60,7 @@ function [priors_cell, trans_mat_cell, em_mat_cell] = estimate_hmm(raw_pr, nn, t
     priors_cell{i} = [prior_off prior_on];
     trans_mat
     em_mat
-    [prior_ff prior on]
+    [prior_off prior_on]
     
   end
 

### 2013-12-11 paper/paper.tex
Worked on the paper, mostly NMF and smoothing
Also small bug fix. I left some modified thresholding code in that I didn't mean to.

@@
 
-Non-negative matrix factorization (NMF) is an unsupervised, dimensionality reduction technique that factors a non-negative matrix $V$ with dimensions $F \times N$ into two matrices,
+Non-negative matrix factorization (NMF) is an unsupervised, dimensionality reduction technique that factors a non-negative matrix $V$ with dimensions $F \times N$ into two non-negative matrices $W$ and $H$.
 
@@
 
-In piano transcription, for example, $V$ might be a spectrogram with dimensions $2048xN$, $W$ might be $2048x88$ and encode the spectral envelope for each piano key in each column of the matrix, and $H$ might be $88xN$ sparse piano roll matrix.
+Typically this factorization is performmed with the goal of recovering some useful structure or information from the matrix factors.
+NMF has been applied to polyphonic transcription and other applications \cite{smaragdis2003non}.
+NMF also has been used for feature extraction for use with other general-purpose classifiers (e.g.
+SVMs) for transcription.
 
-There are two cost functions commonly used for NMF. The first is a euclidean distance cost function, defined as the squared difference between the corresponding elements in each matrix:
+NMF is usually posed as a minimization problem.
+There are two cost functions commonly used for NMF.
+The first is a euclidean distance cost function, defined as the squared difference between the corresponding elements in each matrix:
 
@@
 The functions $||V-WH||^{2}$ and $D(V||WH)$ are convex in W only and H only, but they are not convex in W and H together.
-There is no known procedure for finding the global minimum of these functions, but \citet{seung2001algorithms} have proposed methods for updating W and H such that the measures above will decrease or remain constant after every update.
+There is no known procedure for finding the global minimum of these functions, but \citet{seung2001algorithms} have proposed methods for updating W and H such that the cost functions above will decrease or remain constant after every update.
 Their paper proposes a multiplicative update rule and a gradient descent rule.
@@
 
+\subsection{NMF for Piano Transcription}
+In piano transcription, $V$ might be a spectrogram with dimensions $2048 \times N$, with $N$ corresponding to the length of the song.
+$W$ might be $2048 \times 88$ and encode the spectral envelope for each piano key in each column of the matrix.
+Because of this structure, $W$ is sometimes referred to as a note dictionary matrix.
+$H$ might be $88 \times N$ sparse piano roll matrix.
+It is this strong interpretability of the matrix factors $W$,$H$ that makes NMF an attractive technique for transcription.
+Since $H$ contains the desired note information for transcription, accurately recovering $H$ is the effective goal of NMF for piano transcription.
+After $H$ is recovered, post-processing such as smoothing and thresholding is usually applied to remove spurious activations and improve overall results.
+
 % Applying NMF to the piano transcription problem is not as straightforward as computing the above update rules.
@@
 While a full piano keyboard contains a constant 88 keys, a piano song will often only use a subset of all the keys.
-In practice, pre-selecting the $W$ matrix to find more keys than are present results in the algorithm discovering parts of notes or the same note at different points in the note's ADSR envelope; for example, the attack of a note can have a different spectral signature than its release.
+In practice, pre-selecting the $W$ matrix to find more keys than are present results in the algorithm discovering parts of notes or the same note at different points in the note's Attack-Decay-Sustain-Release (ADSR) envelope; for example, the attack of a note can have a different spectral signature than its release.
 \citet{weninger2013discriminative} use this fact explicitly to learn notes at different points in the ADSR envelope using NMF as a preprocessing step to SVM classification.
 \citet{} uses cross-validation to select the appropriate number of keys.
+\item[$\bullet$] There is no guarentee that a note will have a representation in $W$, even if it is present in the training data.
 \item[$\bullet$] The piano roll matrix $H$ and the spectrogram $V$ are large and often dense matrices, so convergence using the multiplicative update rules can take on the order of minutes on a modern laptop.
-\item[$\bullet$] The notes represented in the $W$ and $H$ matrix are not ordered properly. \citet{} resorted to ordering the notes in $W$ by hand. 
+\item[$\bullet$] The notes represented in the $W$ and $H$ matrix are not ordered properly.
+Ordering does not affect raw performance but greatly detracts from the interpretability.
+One can attempt to estimate the pitch of the columns of $M$ and then sort.
+\citet{abdallah2004polyphonic} found that structuring the initial $W$ would lead to a correctly ordered output.
+\citet{boulangerlewandowski2012} resorted to ordering the notes in $W$ by inspection.
 \end{description}
@@
 
-\subsection{Supervised NMF}
+\subsection{Weakly Supervised NMF}
 
-In the previous NMF methods, we extracted both the note dictionary matrix and the piano roll from the spectrogram.
+In the previous NMF methods, we extracted both the note dictionary matrix $W$ and the piano roll from the spectrogram.
 In piano transcription, however, the dictionary matrix should be similar across multiple transcriptions.
-We can use
+We can guide the NMF process by explicity creating a dictionary matrix $W$ off-line.
+Not only does this alleviate the problem where notes aren't represented in $W$, but it does not require any manual or automatic sorting that is normally required when $W$ is produced by NMF.
+\citet{weninger2013discriminative} suggest that the W matrix may be constructed by performing NMF with rank one (i.e.
+K = 1) on recordings of isolated notes, once for each note.
+The first column of each $W$ matrix is concatenated to form a dictionary matrix for later transcription.
+One can also construct a dictionary matrix by directly concatenating normalized spectral features\cite{niedermayer2008non}.
+
+Some have suggested adding additional bases to $W$.
+For example, one could add multiple examples of the same note played on different pianos to prevent from overemphasizing the characterisics of a specific instrument [cite FINDTHIS].
+\citet{weninger2013discriminative} found including different spectral features for the onset, sustain and decay of each note to be useful.
+In these cases, note activations from different bases but which correspond to the same pitch are summed to form $88 \times N$ piano roll matrix.
+
+Once $W$ is determined, the next step is to recover the piano roll matrix $H$.
+$W$ can be used as an initial value in a standard NMF formulation.
+Another way to recover $H$ is to use the standard multiplicative update rule for $H$ and skip the updating of $W$ since it is already known.
+\citet{niedermayer2008non} refers to this variant of NMF as Non-Negative Matrix \textit{Division}, as we're only estimating $H$ in the approximation.
+
+\subsection{Our Implementation}
+
+We implemented a NMF transcription system.
+We used constructed our $W$ matrix by concatenating spectra from training recording.
+We tried both a standard STFT spectrogram and constant Q transform spectrogram.
+We use the normalizations proposed by \citet{niedermayer2008non}.
+After creating the $W$ matrix, we calculate the $H$ matrix using the previously-defined multiplicative update rule.
+The $H$ matrix is smoothed using a median filter and thresholded.
+[TODO: Results]
 
@@
 \subsection{Smoothing}
+The raw output of many transcription techniques is often noisey and contains many short, spurious notes.
+Additionally, otherwise correctly transcribed notes may contain small unwanted gaps.
+Obviously these detracts from the quality of the transcription.
+Smoothing can be used to improve the transcription by removing the small errors described above.
+Some smoothing techniques are as simple as moving average or median filters, others take advantage of prior knowledge of the temporal structure of music.
+
 \subsubsection{HMM smoothing}

### 2013-12-11 bin/poliner_svm.m
should handle multiple training inputs now

@@ -1,4 +1,4 @@
-function [estimated_pr, pr] = poliner_svm(wav_train, midi_train, wav_test, midi_test)
+function [estimated_pr, pr] = poliner_svm()
   % Warning: this is written for Octave and not tested with MATLAB
   % This also uses libsvm. The libsvm I have included has been built for Octave
 
@@ -16,58 +16,88 @@ function [estimated_pr, pr] = poliner_svm(wav_train, midi_train, wav_test, midi_
   % Testing steps
   % 4. Run SVM on new wave file
   % 5. Run HMM on SVM piano roll output 
-  piano_low = 21;
-  piano_high = 107;
-  more off;
-
-  % Open wav file
-  [y,fs,bps] = wavread(wav_train);
-  [y_2,fs_2,bps_2] = wavread(wav_test);
-  % length(y)
-  % fs
-  % .128*fs
-  %.01*fs
-  % Compute the STFT on 128ms frames with 10ms hops
-  isOctave = exist('OCTAVE_VERSION') ~= 0;
-  if isOctave
-    [S, f, spec_t] = specgram(y, 2048, fs, hanning(.128*fs), 2048 - 160); % Needs to be spectrogram in MATLAB  
-  else
-    [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
-    [S_2, f_2, spec_t_2] = spectrogram(y_2, 2048, 2048-160, 2048, fs_2);
-
-    %[S, f, spec_t] = qgram(y, fs,1); 
-  end
-
-  size(S) 
-  %spec_t(end)
-  % pause
-  % Compute the FFT magnitudes
-  magS = abs(S);
-  magS_2 = abs(S_2);
-  % Normalize, whiten the FFT magnitudes (to-do)
-    
+  
   % Add the libsvm and midi libraries  
   addpath('../lib/libsvm');
   addpath('../lib/matlab-midi/src');
   
-
-  % Open the MIDI file
-  midi = readmidi(midi_train);
-  midi_2 = readmidi(midi_test);
-  notes = midiInfo(midi,0);
-  notes_2 = midiInfo(midi_2,0);
-  %notes = [notes; notes_2];
-  % T is in increments of 10ms, the same as our STFT'd wav file
-  [pr, t, nn] = piano_roll(notes);
-  [pr_2,t_2,nn_2]= piano_roll(notes_2);
-  
+  piano_low = 21;
+  piano_high = 107;
[…]

### 2013-12-11 bin/poliner_svm.m
works

@@ -20,7 +20,7 @@ function [estimated_pr, pr] = poliner_svm()
   % Add the libsvm and midi libraries  
   addpath('../lib/libsvm');
   addpath('../lib/matlab-midi/src');
-  
+  shortener = 3000; %don't grab more than 3000 frames from each piano roll
   piano_low = 21;
   piano_high = 107;
   nn_total = linspace(piano_low,piano_high,87);
@@ -44,13 +44,13 @@ function [estimated_pr, pr] = poliner_svm()
     if isOctave
       [S, f, spec_t] = specgram(y, 2048, fs, hanning(.128*fs), 2048 - 160); % Needs to be spectrogram in MATLAB  
     else
-      [S, f, spec_t] = spectrogram(y, 2048, 2048-160, 2048, fs);
+      [S, f, spec_t] = spectrogram(y(:,1), 2048, 2048-160, 2048, fs);
     end
     magS = abs(S);
-    Spec_trains{i,1}=S;
+    Spec_trains{i,1}=S(:,1:shortener);
     Spec_trains{i,2}=f;
-    Spec_trains{i,3}=spec_t;
-    Spec_trains{i,4}=magS;
+    Spec_trains{i,3}=spec_t(1:shortener);
+    Spec_trains{i,4}=magS(:,1:shortener);
     %Also grab the associated midi file...
     prefix = strsplit(train_name_w,'.');
     prefix = prefix{1};
@@ -66,8 +66,8 @@ function [estimated_pr, pr] = poliner_svm()
     %widen piano rolls to go from 21 to 107
     pr = [zeros(nn(1)-piano_low,size(pr,2));pr];
     pr = [pr;zeros(piano_high-nn(end),size(pr,2))]; 
-    pr_trains{i,1} = pr;
-    pr_trains{i,2} = t;
+    pr_trains{i,1} = pr(:,1:shortener);
+    pr_trains{i,2} = t(:,1:shortener);
   end  
     
   for i=1:size(test_wavs,1)
@@ -99,7 +99,7 @@ function [estimated_pr, pr] = poliner_svm()
   end    
    
   %notes
-  pause
+  
   
   
   size(pr)
@@ -107,7 +107,7 @@ function [estimated_pr, pr] = poliner_svm()
   % sum(pr,2)
 
   % Only display a subset of the ground truth for comparison to computed piano roll later
-  subset = 1000;
+  subset = 2000;
[…]

### 2013-12-11 paper/paper.tex
Added a song table like Poliner to the paper and added dictionary figures

@@
 \providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
+
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
@@
 
+\begin{figure*}[t]
+\begin{center}
+\subfigure[]{
+\includegraphics[scale=.6]{stft-dictionary}
+\label{fig:stftdict}
+}
+\subfigure[]{
+\includegraphics[scale=.6]{cqt-dictionary}
+\label{fig:cqtdict}
+}
+\end{center}
+\caption{(a) shows the W dictionary matrix computed for the supervised NMF method.
+Each column represents the spectral signature of a piano note.
+Notice how the base frequencies follow an exponential curve.
+(b) shows the H piano roll matrix output from the supervised NMF method.}
+\end{figure*}
+
 \subsection{Non-negative matrix factorization}
@@
 \begin{description}
-\item[$\bullet$ Jeremy] wrote and evaluted the SVM method in \citet{poliner2006discriminative}, the Bayesian nonparametric method in \citet{blei2010bayesian}, and the sparse non-negative matrix factorization method in \citet{abdallah2004polyphonic}.
-Jeremy also wrote the paper.
-\item[$\bullet$ Mark] implemented the hidden Markov model in \citet{poliner2006discriminative} and the LSTM network in \citet{bock2012polyphonic}.
-\item[$\bullet$ Paul] implemented the LSTM network in \citet{bock2012polyphonic}.
+\item[$\bullet$ Jeremy] wrote and evaluted the SVM method in \citet{poliner2006discriminative}, the non-negative matrix factorization method in \citet{abdallah2004polyphonic}, the AdaBoost method in \citet{boogaart2009note}, and wrote the paper.
+\item[$\bullet$ Mark] implemented the hidden Markov model and error measures in \citet{poliner2006discriminative}, evaluated the methods, and wrote the paper.
+\item[$\bullet$ Paul] implemented the non-negative matrix factorization method in \citet{abdallah2004polyphonic} and wrote the paper.
 \end{description}
+
+\begin{center}
+\begin{table*}[ht]
+%   \caption{Table of midi songs used for testing and training the algorithms in the paper.}
+\caption{MIDI songs used for testing and training the methods in the paper.}
+\begin{center}
+% \setlength{\tabcolsep}{4.0em}
+\begin{tabular}{L{2cm} L{4cm} L{4cm}}
+% \textbf{Composer} & \textbf{Training} & \textbf{Testing} \\ \hline
+Composer & Training & Testing \\ \hline
+Brahms & 
+Sonata C major, Op. 1 (Allegro, Andante), 7 Fantasia, Op. 116 (Andante), 3 Intermezzi, Op. 117 (Andante moderato)
+& 
+Sonata C major, Op. 1 (Finale), 7 Fantasia, Op. 116 (Andantino teneramente), 3 Intermezzi, Op. 117 (Andante moderato)
+\\
+
+Beethoven & Sonata No. 5, Op. 10 (Prestissimo), Sonata No. 8, Op. 13 (Rondo Allegro) & Sonata No. 11, Op. 22 (Minuetto), Sonata No. 21, Op. 53 (Adagio molto) \\
+
+Chopin & 5 Mazurkas, Op. 7 (Vivace), Etudes, Op. 10 (Allegro, Vivace), Etudes, Op. 25 (Allegro) 
+& 5 Mazurkas, Op. 7 (Vivo ma non troppo), Etudes, Op. 10 (Allegro con fuoco), Etudes, Op. 25 (Lento)
+\\
+Debussy & Suite bergamasque (Moderato, Menuet) & Suite bergamasque (Andante, Allegretto) \\
+Mozart & Sonata No. 8 (Andantino), Sonata No. 12 (Adagio)
+ & Sonata No. 8 (Allegro con spirito), Sonata No. 16 (Allegro)  \\
+
+Schubert & 
+6 Moments musicaux, Op. 94 (Allegretto moderato), Fantasia C major (Presto)
+& 6 Moments musicaux, Op. 94 (Allegro vivace), Fantasia C major (Allegro)
+\\
+
+% AdaBoost & 22\% & 22 \\ 
+\end{tabular}
+\end{center}
+\end{table*}
+\end{center}
 

### 2013-12-11 paper/paper.tex
Fleshed out the piano physics and adaboost sections

@@
 
-\citet{bank2003physically} details the physics of piano sound synthesis. We will just note here that...
+The physical description of sound production in a piano is complex.
+\citet{bank2003physically} describes physics-based synthesis algorithm that models the piano hammer as a nonlinear spring connect to a lumped mass, a nonideal, lossy piano string with digital waveguides, and soundboard radiation with a linear filter.
+This model gives rise to the rich and variable set of overtones across the keyboard.
+Furthermore, in a procedure called stretched tuning, real pianos have their higher notes tuned slightly higher and their lower notes tuned slightly lower to compensate for psychoacoustic detuning effect in our ears.
+These factors give rise to the complex spectral feature space for our machine learning methods.
 
-Figure X shows the spectral profile of a low C note played over a couple seconds.
-
-This is typically the lowest 
-(But it's more complicated than that...)
-- Show why it's complicated 
-- Show some of the physics
-
-\section{Related Work}
+% \section{Related Work}
 % Contain a table with metrics data, brief description, year published
@@
 
-\citet{boogaart2009note} proposes a cascade of boosted classifiers.
+\citet{boogaart2009note} proposes using AdaBoost with decision stumps for the weak learners to detect note onsets.
+The motivation for using AdaBoost in transcription over other methods is its robustness to overfitting the spectral features and its interpretability in the transcription process.
+The interpretation is that the decision stumps decide on the presence of a note based on the magnitudes of the fundamental frequency of a note and then its overtones.
+In some sense, this is a more sophisticated version of applying a threshold to the constant Q transform, but it suffers from the same problems; the method has difficulty with volume changes in the song, and dense chords with many overlapping overtones can also confuse the decision stumps.
 
-\subsection{Feed forward neural network}
+% \subsection{Feed forward neural network}
 
@@
 This database, along with the MAPS database, are a popular choice for piano transcription researchers.
-Table X gives a list of the composers and songs we used for testing.
+Table \ref{table:songs} on page 7 gives a list of the composers and songs we used for testing and training.
 
@@
 \end{center}
+\label{table:songs}
 \end{table*}

### 2013-12-11 paper/paper.tex
Fleshed out more sections of the paper

@@
 The first approach learns a supervised learning model that uses frequency magnitudes to predict the presence of a note.
-The second approach uses unsupervised learning techniques to extract
+The second approach uses unsupervised learning techniques to simultaneously extract notes and a musical score from a spectrogram.
+The following sections detail the most popular techniques within these general approaches.
 
-\subsection{Audio data}
+%\subsection{Audio data}
 
-Our training data comes from a large database of piano MIDI files. MIDI is a standard for specifying...
+%Our training data comes from a large database of piano MIDI files. MIDI is a standard for specifying...
 %musical score
@@
 This is a simpler problem than detecting note durations because of the possibility for time-varying timbres in instruments.
-Most researchers in this field evaluate their music transcription methods on their note onset detection performance.
+Many researchers in this field evaluate their music transcription methods on their note onset detection performance.
+In this paper, we chose to evaluate our methods over the entire duration of the notes; while this brings us closer to the original musical score, it also results in higher error measures.
 
@@
 % SVM, logistic regression
+Applying discriminative machine learning models to detect notes is a dominant approach in transcription research.
+A discriminative model can be learned for each note (for pianos, 88 classifiers) over the frequency magnitudes of the song's spectrogram.
+Each note classifier outputs whether the note is present or absent at each time step determined by the time step used in the spectrogram.
+Many researchers have chosen nonlinear or kernel-based methods due to the complexity of the input space.
+
 \subsection{SVM}
 
-% one vs all
 \citet{poliner2006discriminative}
 
+% use support vector machines to predict note presence or absence at 10ms time steps based on
+
+% one vs all
 % frequency partition
@@
 
-\citet{nam2011classification} proposes ...
- training a restricted Boltzmann machine on the spectrogram features. The features computed by the RBM are used by an SVM
+To increase the previous system's robustness to noise, \citet{nam2011classification} proposed a preprocessing stage using a restricted Boltzmann machine on the spectrogram features.
+The RBM would be able to learn noise invariant features over the spectrogram and provide a new input to the SVM classifiers.
+The researchers found a modest improvement in performance as they increased noise levels in their input data.
 
@@
 \subsection{Bayesian nonparametric models}
-One of the downsides of the NMF methods is that they require the number of notes (or sources) to be specified
+One of the downsides of the NMF methods is that they require the number of notes (or sources) to be specified in advance.
+In piano transcription, we know that there are a constant 88 keys, but it is very often the case that only a subset of all keys will be used in a given song.
+Choosing the number of sources to be 88 in this case often results in duplicate notes or parts of notes in the dictionary matrix.
+\citet{blei2010bayesian} offers a Bayesian nonparametric method for transcription, similar to topic modeling, that automatically determines the appropriate number of sources.
 
 \subsection{Smoothing}
-The raw output of many transcription techniques is often noisey and contains many short, spurious notes.
+One downside of the previous methods is that each time step and each note is classified independently of every other note.
+The raw output of these techniques is often noisy and contains many short, spurious notes.
 Additionally, otherwise correctly transcribed notes may contain small unwanted gaps.
-Obviously these detracts from the quality of the transcription.
+These gaps detract from the quality of the transcription.
 Smoothing can be used to improve the transcription by removing the small errors described above.
@@
 
-\subsubsection{Probabilistic spectral smoothness}
-
-\subsection{Recurrent neural network}
+%\subsubsection{Probabilistic spectral smoothness}
+%\subsection{Recurrent neural network}
 
@@
 
-The MIDI music used to evaluate these methods came from the Classical Piano MIDI Page at http://www.piano-midi.de/, a free, online database of piano MIDI files with a wide range of composers.
+The MIDI music we used to evaluate these methods comes from the Classical Piano MIDI Page\footnote{http://www.piano-midi.de/}, a free, online database of piano MIDI files with a wide range of composers.
 This database, along with the MAPS database, are a popular choice for piano transcription researchers.
@@
 
-The MIDI files represent the ground truth data. To generate training data, we synthesized wav files from the MIDI files using Timidity...
-
-Synthesized with Timidity -> 16kHz
+The MIDI files represent the ground truth data and the training labels for our classifiers.
+To generate training features, we synthesized the MIDI files into WAV files using Timidity++\footnote{http://timidity.sourceforge.net/}.
+Since the highest fundamental frequency on a piano is 4186Hz, we chose a sampling rate of 16kHz to capture the majority of the overtone frequencies.
+Many researchers choose an 8kHz sampling rate to decrease the size of their input data.
 
@@
 
-There is no standard metric to evaluate the performance of a transcription method. However, many researchers use the following methods:
+There is no standard metric to evaluate the performance of a transcription method. Researchers use a variety of error measures, described in the sections below.
 
@@
 \section{Conclusion}
+
+We achieved the best results with...
+
 % Which method is fastest

### 2013-12-11 paper/paper.tex
Wrote about audio data. Edited a caption

--- 2013-12-11 paper/paper.tex
@@
 \subsection{Audio data}
+Our training data comes from a large database of piano MIDI files.
+MIDI is a standard for specifying the controlling and playing of digital instruments.
+A MIDI specifies the instruments, notes, volumes and the other information needed to digitally synthesize(sp?) a song.
+MIDI files may specify a type of instrument but the exact sonic characteristics of the virtual instrument are defined by the synthesizer.
+That is, two identical midi files played with different sythnesizers may sound different.
 
-Our training data comes from a large database of piano MIDI files. MIDI is a standard for specifying...
+MIDI is an eminently useful technology for the development of transcription technology.
+It contains both the time-coded note information and is capable of producing the corresponding audio file, which are respectively the output and input of the transcription system.
+Having raw audio and note information that are time-aligned is important as the note information can act as the labels in supervised learning.
+
+Digital audio can be stored using WAV files.
+WAV files are uncompressed and directly store digital audio samples.
+A typical WAV file is at CD quality: 16 bit samples at 44.1 KSa/s.
+We found that many transcription techniques downsample the audio (e.g.
+to 8 or 16 KSa/s) before processing, with the idea being that the higher frequency information lost in downsampling is not useful for transcription and that transcription is faster with fewer samples to process \citet{}.
 %musical score
@@
 \end{center}
-\caption{(a) shows the W dictionary matrix computed for the supervised NMF method.
+\caption{ Two W dictionary matrices computed for the supervised NMF method.
 Each column represents the spectral signature of a piano note.
-Notice how the base frequencies follow an exponential curve.
-(b) shows the H piano roll matrix output from the supervised NMF method.}
+(a) The W dictionary matrix computed using an STFT.
+(b) The W dictionary matrix computed using constant Q transform.}
 \end{figure*}

### 2013-12-11 paper/paper.tex

--- 2013-12-11 paper/paper.tex
@@
 
-%\subsection{Audio data}
+\subsection{Audio data}
+Our training data comes from a large database of piano MIDI files.
+MIDI is a standard for specifying the controlling and playing of digital instruments.
+A MIDI specifies the instruments, notes, volumes and the other information needed to digitally synthesize(sp?) a song.
+MIDI files may specify a type of instrument but the exact sonic characteristics of the virtual instrument are defined by the synthesizer.
+That is, two identical midi files played with different sythnesizers may sound different.
 
-%Our training data comes from a large database of piano MIDI files. MIDI is a standard for specifying...
+MIDI is an eminently useful technology for the development of transcription technology.
+It contains both the time-coded note information and is capable of producing the corresponding audio file, which are respectively the output and input of the transcription system.
+Having raw audio and note information that are time-aligned is important as the note information can act as the labels in supervised learning.
+
+Digital audio can be stored using WAV files.
+WAV files are uncompressed and directly store digital audio samples.
+A typical WAV file is at CD quality: 16 bit samples at 44.1 KSa/s.
+We found that many transcription techniques downsample the audio (e.g.
+to 8 or 16 KSa/s) before processing, with the idea being that the higher frequency information lost in downsampling is not useful for transcription and that transcription is faster with fewer samples to process \citet{}.
 %musical score

### 2013-12-11 paper/paper.tex
added stuff to intro, data

@@
 % This is just copied from 'A Connectionist Approach to Automatic Transcription of Polyphonic Piano Music' for now:
-Applications of a music transcription are versatile.
-Transcription produces a compact and standardized parametric representation of music.
-Such representation is needed for content-based retrieval of music in most current musical databases.
-It is useful in music analysis systems for tasks such as melody extraction, music segmentation and rhythm tracking.
+Music transcription is the process of creating a musical score from an audio recording.
+There are many applications of music transcription.
+Transcription produces a compact and standardized parametric representation of music, which is necessary for content-based retrieval of music in most current musical databases.
+It is also useful in music analysis systems for tasks such as melody extraction, music segmentation and rhythm tracking.
 Transcription aids musicologists in analyzing music that has never been written down, such as improvised or ethnical music.
 The conversion of an acoustical waveform into parametric description is also useful in the process of making music, as well as in newer coding standards, such as MPEG-4, which may include such descriptions.
+
+Historically, the task of music transcription has been left to expert musicians.
+However, the process is difficult and time-consuming for complex recordings.
+Thus the ability to automatically transcribe music is of both practical and theoretical interest.
 
@@
 \subsubsection{HMM smoothing}
+\begin{figure*}[t]
+\begin{center}
+\subfigure[]{
+\includegraphics[scale=.6]{raw_svm}
+}
+\subfigure[]{
+\includegraphics[scale=.6]{smoothed_svm}
+}
+\end{center}
+\caption{ Many of the classifiers do not take temporal structure into account.
+HMM smoothing captures some of this structure.
+(a)  The estimated piano-roll before HMM smoothing.
+(b)  The estimated piano-roll after HMM smoothing}
+\end{figure*}
 In many of the techniques used (such as SVM's), notes at a single time step are classified independently of the notes at adjacent time steps.
@@
 \hline
-Algorithm & Accuracy & Score \\ \hline
-SVM & 21\% & 21 \\ \hline
-SVM with HMM & 22\% & 22 \\ \hline
+Algorithm & Accuracy & $E_{tot}$ \\ \hline
+SVM & 67.7\% & NR \\ \hline
+SVM with HMM & 70\% & 34.2 \\ \hline
 NMF & 23\% & 23 \\ \hline
@@
 \hline
-Algorithm & Accuracy & Score \\ \hline
-SVM & 18\% & 18 \\ \hline
-SVM with HMM & 19\% & 19 \\ \hline
+Algorithm & Accuracy & $E_{tot}$ \\ \hline
+SVM & 51.7\% & 84.7  \\ \hline
+SVM with HMM & 62.7\% & 53.8  \\ \hline
 NMF & 20\% & 20 \\ \hline
@@
 
-Beethoven & Sonata No. 5, Op. 10 (Prestissimo), Sonata No. 8, Op. 13 (Rondo Allegro) & Sonata No. 11, Op. 22 (Minuetto), Sonata No. 21, Op. 53 (Adagio molto) \\
+Beethoven & Sonata No. 5, Op. 10 (Prestissimo), Sonata No. 8, Op. 13 (Rondo Allegro), For Elise (Poco moto) & Sonata No. 11, Op. 22 (Minuetto), Sonata No. 21, Op. 53 (Adagio molto) \\
 

### 2013-12-11 paper/paper.tex
added svm section
Added some parts to the paper

@@
 
-\citet{poliner2006discriminative}
-
 % use support vector machines to predict note presence or absence at 10ms time steps based on
-
+\citet{poliner2006discriminative} first used short-time Fourier transforms were applied to the audio data.
+Then,  87(they omit the highest note) one-versus-all (OVA) SVM classifiers were trained, one for each note.
+The SVM's were then used to predict the test data, and HMM's were used for post-process smoothing.
 % one vs all
@@
 This sparse NMF method in \citet{hoyer2004non} was adopted by \citet{abdallah2004polyphonic} in their transcription system.
-
 % Non-negative Matrix factorization
@@
 
+% \subsubsection{Median filtering}
+
 \section{Evaluation}
@@
 Algorithm & Accuracy & $E_{tot}$ \\ \hline
-SVM & 67.7\% & NR \\ \hline
+SVM & 67.7\% & - \\ \hline
 SVM with HMM & 70\% & 34.2 \\ \hline
-NMF & 23\% & 23 \\ \hline
-Sparse NMF & 24\% & 24 \\ \hline
-AdaBoost & 25\% & 25 \\ \hline
+Sup. NMF & - & - \\ \hline
+AdaBoost & 75.2\% & - \\ \hline
 \end{tabular}
 \end{center}
+% precision, recall, f
+% 42.3, 68.8, .524
 
@@
 SVM with HMM & 62.7\% & 53.8  \\ \hline
-NMF & 20\% & 20 \\ \hline
-Sparse NMF & 21\% & 21 \\ \hline
-AdaBoost & 22\% & 22 \\ \hline
+Sup. NMF & 30.4\% & 92.4 \\ \hline
+AdaBoost & 42.5\% & 113.9 \\ \hline
 \end{tabular}

### 2013-12-11 paper/paper.tex
Added figures and stuff to the paper

--- 2013-12-11 paper/paper.tex
@@
 
-Polyphony means multiple simultaneous notes. Monophony means only a single note at a time. These terms carry slightly different meanings outside of automated transcription research.
+Polyphony means multiple simultaneous notes. Mono-phony means only a single note at a time. These terms carry slightly different meanings outside of automated transcription research.
 
@@
 MIDI is a standard for specifying the controlling and playing of digital instruments.
-A MIDI specifies the instruments, notes, volumes and the other information needed to digitally synthesize(sp?) a song.
+A MIDI specifies the instruments, notes, volumes and the other information needed to digitally synthesize a song.
 MIDI files may specify a type of instrument but the exact sonic characteristics of the virtual instrument are defined by the synthesizer.
@@
 \subsection{SVM}
+% use support vector machines to predict note presence or absence at 10ms time steps based on
 
-% use support vector machines to predict note presence or absence at 10ms time steps based on
-\citet{poliner2006discriminative} first used short-time Fourier transforms were applied to the audio data.
-Then,  87(they omit the highest note) one-versus-all (OVA) SVM classifiers were trained, one for each note.
-The SVM's were then used to predict the test data, and HMM's were used for post-process smoothing.
+\citet{poliner2006discriminative} first proposed applying support vector machines to piano transcription.
+Their system uses 87 SVM classifiers, one for each piano key (but omitting the highest piano key), that predicts if its note is present or absent given the frequency magnitudes at a given time step.
+The SVMs are trained in a one-versus-all fashion.
+To speed up training and testing, the input space of frequency magnitudes is partitioned into lower and higher frequencies for the lower note and higher note classifiers, respectively.
+Their paper also implements a hidden Markov model to smooth the raw outputs from the SVMs.
+
 % one vs all
@@
 \subfigure[]{
-\includegraphics[scale=.6]{cqt-dictionary}
-\label{fig:cqtdict}
+\includegraphics[scale=.6]{H-matrix}
+\label{fig:H-matrix}
 }
@@
 The $H$ matrix is smoothed using a median filter and thresholded.
-[TODO: Results]
 
@@
 
-The following table captures our evaluation of these methods.
+Table \ref{table:ourevals} captures our evaluation of these methods.
 We expect these accuracy measures to be lower than they were reported in the original papers because we are using a smaller training dataset and not attempting to optimize the methods.
@@
 \begin{center}
+\label{table:ourevals}
 \begin{tabular}{|l|l|l|}
 \hline
+
 Algorithm & Accuracy & $E_{tot}$ \\ \hline
@@
 AdaBoost & 42.5\% & 113.9 \\ \hline
+
 \end{tabular}
+
 \end{center}
@@
 
-We achieved the best results with...
+We achieved the best accuracy with the SVM method with HMM smoothing from \citet{poliner2006discriminative}.
 

### 2013-12-11 paper/paper.tex
Made nmf more testable, some paper edits

@@
 The $H$ matrix is smoothed using a median filter and thresholded.
+The STFT $W$ matrix and a sample of the smoothed $H$ are shown in figure 2.
 
@@
 \item[$\bullet$ Mark] implemented the hidden Markov model and error measures in \citet{poliner2006discriminative}, evaluated the methods, and wrote the paper.
-\item[$\bullet$ Paul] implemented the non-negative matrix factorization method in \citet{abdallah2004polyphonic} and wrote the paper.
+\item[$\bullet$ Paul] implemented the non-negative matrix factorization methods in \citet{abdallah2004polyphonic} and \citet{niedermayer2008non} and wrote the paper.
 \end{description}

### 2013-12-11 bin/midi_cache.m

@@ -1,6 +1,6 @@
 function [pr, nn] = midi_cache(midi_file)
   more off;
-  mat_cache = strcat(midi_file, '.cache');
+  mat_cache = strcat(midi_file, '.cache.mat');
   try 
       load(mat_cache);
       fprintf('Cached MIDI found!\n');

### 2013-12-11 bin/nmf_train.m

@@ -1,0 +1,89 @@
+function [ W ] = nmf_train(use_qgram, draw_plots)
+
+
+  % This also uses libsvm. The libsvm I have included has been built for Octave
+  more off;
+if (nargin < 1)
+    use_qgram = 0;
+    draw_plots = 0;
+end
+
+if (nargin == 1)
+    draw_plots = 0;
+end
+    
+  % Get list of training notes
+isOctave = exist('OCTAVE_VERSION') ~= 0;
+if isOctave
+     filelist = readdir('../data/notes');
+  for i = 1:numel(filelist)
+    if (regexp (filelist{i},'^\w+.wav$'))
+        wavs{i} = filelist{i};
+    end
+  end
+else
+  filelist = dir('../data/notes');
+  count = 1;
+  for i = 1:numel(filelist)
+    if (regexp (filelist(i).name,'.wav$'))
+        wavs{count} = filelist(i).name;
+        count = count + 1;
+    end
+  end
+end
+
+bins = 2049;
+if (use_qgram)
+    bins = 175;
+end
+W = zeros(bins, 88);
+for i = 1:numel(wavs)
+    if (~isempty(wavs{i}))
+        
+      % Read File
+      wavs{i} = strcat('../data/notes/', wavs{i});
+      disp(wavs{i});      % Constant Q Transform
+      if (use_qgram)
+        [S, f, spec_t] = qgram_cache(wavs{i}); 
+      else
+        [y,fs,bps] = wavread(wavs{i});
+        y = [zeros(length(y),1); y];
+        [S, f, spec_t] = spectrogram(y, 4096, 4096-160, 4096, fs);    
+      end
+
+      % Crop low energy spectra
+      S = abs(S);
+      Sen = (sum(S).^2);
+      thresh = 100;
+      if (use_qgram)
+          thresh = 1e-5;
+      end
+          
+      S(:, Sen < thresh) = 0;
+
+      % Average the signal-containing spectra
+      Sm = mean(S, 2);
+
+      % Magnitude warping and mean subtraction proposed by Niedermeyer
+      g = mean(Sm); % What average should we use (mean? outlier resistant?)
+      S_warp = log(1 + (1/g).*Sm);
+
+      %%y = smooth(S_warp, 9);
+      y = filter(ones(9,1)/9, 1, S_warp);
+      wi = max(0, S_warp - y);
+
+      if (draw_plots)
+          plot(wi);
[…]

### 2013-12-11 paper/paper.tex
added stuff

--- 2013-12-11 paper/paper.tex
@@
 
-\section{Introduction}
+\section{Terminology}
 
-\subsection{Terminology}
 Semitones are the smallest music interval in Western music. On a piano, semitones are the adjacent keys.
@@
 Some have suggested adding additional bases to $W$.
-For example, one could add multiple examples of the same note played on different pianos to prevent from overemphasizing the characterisics of a specific instrument [cite FINDTHIS].
+For example, one could add multiple examples of the same note played on different pianos to prevent from overemphasizing the characterisics of a specific instrument.
 \citet{weninger2013discriminative} found including different spectral features for the onset, sustain and decay of each note to be useful.
@@
 \subsubsection{Reported measures}
+The table below reports the error measures from the original papers.
+$E_{tot}$, originally defined in \citet{poliner2006discriminative}, does not appear in all the papers, and the supervised NVM method only reports precision, recall, and f-score figures.
+
 \begin{center}
@@
 SVM with HMM & 70\% & 34.2 \\ \hline
-Sup. NMF & - & - \\ \hline
+Sup. NMF & 32.3\% & - \\ \hline
 AdaBoost & 75.2\% & - \\ \hline
@@
 Table \ref{table:ourevals} captures our evaluation of these methods.
-We expect these accuracy measures to be lower than they were reported in the original papers because we are using a smaller training dataset and not attempting to optimize the methods.
+We expect these accuracy measures to be lower than they were reported in the original papers because we are using a smaller training dataset and left the methods unoptimized.
 
@@
 SVM with HMM & 62.7\% & 53.8  \\ \hline
-Sup. NMF & 30.4\% & 92.4 \\ \hline
+Sup. NMF & 30.8\% & 151.0 \\ \hline
 AdaBoost & 42.5\% & 113.9 \\ \hline
@@
 We achieved the best accuracy with the SVM method with HMM smoothing from \citet{poliner2006discriminative}.
+Across all methods, the raw output contained lots of spurious misclassifications, and smoothing was crucial for increasing accuracy, whether by hidden Markov models in the SVM method or median filtering with supervised NMF and AdaBoost.
+The AdaBoost method had suprisingly low accuracy relative to the other methods, although we spent the least amount of time optimizing this method.
+
+We were surprised by the difficulties automated transcription presents.
+The ASDR envelope of piano notes complicates NMF retrieval of individual notes and results in duplicated notes.
+Cross-validating smoothing parameters was important for better accuracy.
+Aligning the spectrogram with ground truth data becomes nontrivial with the constant Q transform; unlike an FFT that has a definable center, the window size of the constant Q transform changes with each frequency.
 
 % Which method is fastest
-
 % Which method has the best performance
-
 % Which one is the simplest method
-
 % Future directions
